### 2024-11-26 other/datasheet/datasheet.qmd
update paper structure

---
title: "Datasheet for Wages by Education Level"
author: 
  - Yuanyi (Leo) Liu
thanks: "Code and data are available at: https://github.com/leoyliu/Analyzing-the-Impact-of-Education-on-Earnings-in-Canada"
date: today
date-format: long
abstract: "This datasheet describes the \"Wages by Education Level\" dataset, which records hourly wage rates across different levels of education in Canada from 2000 to 2019. It focuses on individuals aged 25-54 and aims to facilitate analysis of how education impacts wages. The dataset is available under the Open Government Licence - Ontario, allowing for broad use in research and policy discussions. It is maintained by the Government of Canada, ensuring its accuracy and applicability for continuous economic studies."
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from [@gebru2021datasheets]. The data is obtained from the Government of Ontario [@citedata].

**Motivation**

1.  *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    -   The dataset was constructed to examine the relationship between education level and average hourly wage rates in Canada between the years 2000 and 2019. The intent was to analyze whether educational attainment translates into higher earnings and how this has evolved over time.
2.  *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    -   Yuanyi (Leo) Liu, while studying at the University of Toronto?
3.  *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    -   No applicable.
4.  *Any other comments?*
    -   No additional comments are provided.

**Composition**

1.  *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
    -   The dataset represents aggregated wage data by educational attainment for individuals within the age group of 25-54 in Canada. It does not contain multiple types of instances.
2.  *How many instances are there in total (of each type, if appropriate)?*
    -   There are 180 instances in the dataset, each representing a unique combination of year and education level.
3.  *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
    -   The dataset appears to be a complete set of instances from the available data during the specified timeframe. Representativeness is assumed but not specifically validated.

4.  *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
    -   Each instance consists of processed data featuring the year, the level of education, age group, and the average hourly wage rate for the corresponding education level.

5.  *Is there a label or target associated with each instance? If so, please provide a description.*
    -   The average hourly wage rate serves as the primary variable of interest, which could be considered the target for analyses examining wage differences across education levels.

6.  *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
    -   There is no information indicated as missing from the dataset.

7.  *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
    -   Relationships are not explicitly stated but can be inferred through the year and education level variables.

8.  *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
    -   There are no recommended data splits mentioned, as the data appears to be used for comprehensive temporal analysis rather than predictive modeling that would require such splits.

9.  *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
    -   No specific errors, noise, or redundancies are noted. However, the potential for such issues exists as with any dataset and would be subject to further validation checks.

10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
    -   The dataset is self-contained and does not rely on external resources.

11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
    -   The dataset does not contain confidential information as it is aggregated data that does not identify individuals.

12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
    -   The dataset does not contain any content that might be offensive or otherwise cause anxiety.

13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
    -   The dataset identifies sub-populations based on education level but does not provide further demographic breakdowns such as age, gender, etc.

14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
    -   It is not possible to identify individuals from the dataset as the data is aggregated and anonymized.

15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
    -   The dataset does not contain data that might be considered sensitive as it deals with aggregated economic figures without any direct personal identifiers.

16. *Any other comments?*
    -   No additional comments are provided.

**Collection process**

1.  *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
    -   The data associated with each instance was acquired from a governmental database, ensuring that the data was directly observable and recorded through standardized bureaucratic processes. Given that the data originates from official government records, there is a high degree of validation inherent in the data collection method.

2.  *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
    -   The data was collected using government administrative procedures and mechanisms. These include labor surveys and economic reports which are routinely validated through governmental audit and review processes.

3.  *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
    -   As the dataset was derived from a comprehensive administrative record, the sampling strategy is not applicable. The data represents a census of the information available rather than a sample.

4.  *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
    -   The collection of the data was performed by government employees as part of their standard duties. Compensation follows the governmental pay scale for such positions.

5.  *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
    -   The data was collected annually, reflecting the earnings within that year, thus ensuring that the timeframe of data collection matches the creation timeframe of the data instances.

6.  *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
    -   The data being from a government source and aggregated for public use, individual ethical review processes are typically not required beyond the government's standard data collection and dissemination policies.

7.  *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
    -   The data was obtained directly from government sources and not through third parties, ensuring direct responsibility and accountability for the data accuracy.

8.  *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
    -   Notification for data collection are not typically required for governmental data collection at this aggregate level, as it is a standard practice for economic and labor data reporting and does not involve individual-level data.

9.  *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
    -   Consent for data collection are not typically required for governmental data collection at this aggregate level, as it is a standard practice for economic and labor data reporting and does not involve individual-level data.

10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
    -   Given the nature of the data as aggregate and anonymized, mechanisms for individuals to revoke consent are not applicable.

11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
    -   The potential impact of the dataset and its use on data subjects is generally considered in the context of privacy and data protection laws under which the data is collected and released by the government. However, specific documentation of impact analysis would be under the purview of the governmental body responsible for the data.

12. *Any other comments?*
    -   No additional comments are provided.

**Preprocessing/cleaning/labeling**

1.  *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
    -   Preprocessing and cleaning were applied to the dataset, which included filtering specific subsets of data relevant to the study, such as data pertaining to the Canadian geography, average hourly wage rate, and specific age groups. The education levels were also ordered numerically to facilitate analysis. Missing values or incomplete records were not reported; however, any that may have been present were likely excluded during the filtering process.

2.  *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
    -   The raw data remains stored and is accessible as part of the government's open data portal. This ensures transparency and allows for replication or alternate analyses by other researchers.

3.  *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
    -   The R scripts used for preprocessing, cleaning, and labeling are available within the project's repository. They provide a transparent methodological framework for replicating the dataset processing steps.

4.  *Any other comments?*
    -   No additional comments are provided.

**Uses**

1.  *Has the dataset been used for any tasks already? If so, please provide a description.*
    -   The dataset has been used for an econometric analysis to understand the relationship between education level and average hourly wage rate in Canada within the specified age group and time frame.

2.  *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
    -   https://github.com/leoyliu/Analyzing-the-Impact-of-Education-on-Earnings-in-Canada

3.  *What (other) tasks could the dataset be used for?*
    -   Beyond the current econometric analysis, the dataset could be used for a range of studies in labor economics, education policy, wage disparity analysis, and longitudinal studies on wage progression over time.

4.  *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
    -   The way the dataset has been collected and processed should be considered when conducting future research to ensure that the socioeconomic context is taken into account, especially considering the dataset only covers a specific age group and may not generalize across all demographics. Future users should ensure that their research questions and methods are compatible with the dataset's limitations to avoid skewed interpretations.

5.  *Are there tasks for which the dataset should not be used? If so, please provide a description.*
    -   The dataset is not suited for analysis that requires individual-level detail or for making assumptions outside the scope of the variables provided, such as inferring reasons behind wage differences without considering other socioeconomic factors.

6.  *Any other comments?*
    -   No additional comments are provided.

**Distribution**

1.  *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
    -   The dataset will not be directly distributed by the creators but is available through the Open Government Canada platform. This ensures that the dataset remains accessible to the public and researchers without any restrictions imposed by the original creators.

2.  *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
    -   The dataset is accessible via a direct download link provided by Open Government Canada. There is no digital object identifier (DOI) specifically assigned to this dataset.

3.  *When will the dataset be distributed?*
    -   The dataset is already available for public access and has been since its inclusion in the government's open data portal.

4.  *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
    -   The dataset is distributed under the Open Government Licence – Ontario, which allows for free use and redistribution of the data with attribution. More details on the licensing terms can be found on the Open Government Canada website.

5.  *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
    -   There are no known third-party IP restrictions on this dataset, as it is provided by a governmental body with the intent for public use.

6.  *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
    -   No export controls or other regulatory restrictions are known to apply to this dataset. It is intended for public use within the legal frameworks governing public data in Canada.

7.  *Any other comments?*
    -   No additional comments are provided.

**Maintenance**

1.  *Who will be supporting/hosting/maintaining the dataset?*
    -   The dataset is maintained by the Government of Canada, specifically through its Open Government initiative.

2.  *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
    -   The dataset can be inquired about through the contact mechanisms provided on the Open Government Canada platform.

3.  *Is there an erratum? If so, please provide a link or other access point.*
    -   Any corrections or updates to the dataset are handled by the Open Government Canada team and communicated through updates on the platform where the dataset is hosted.

4.  *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
    -   The dataset is updated periodically as new wage data becomes available. These updates are documented and communicated through the Open Government Canada platform.

5.  *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
    -   As the data is aggregated and anonymized, there are no specific retention limits applicable to this dataset. The data is intended for long-term use as part of governmental economic studies.

6.  *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
    -   Older versions of the dataset are archived and remain accessible via the Open Government Canada platform to ensure that historical data is available for longitudinal studies.

7.  *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
    -   There are currently no mechanisms for external contributions to directly augment or extend the dataset. It is maintained exclusively by governmental statisticians and data clerks.

8.  *Any other comments?*
    -   No additional comments are provided.

\newpage

# References

### 2024-11-26 paper/paper.qmd
update paper structure

---
title: "Determinants of Wage Variation in Canada"
subtitle: "The Influence of Education, Gender, and Age on Canadian Earnings"
author: 
  - Yuanyi (Leo) Liu
thanks: "Code and data are available at: [Determinants of Wage Variation in Canada](https://github.com/leoyliu/Determinants-of-Wage-Variation-in-Canada)"
date: today
date-format: long
abstract: "In this paper, we explore the connection between education and hourly wages in Canada, focusing on data from the year 2000. Our findings indicate a clear trend: higher educational levels correlate with increased average hourly wages for individuals aged 25 to 54. This research highlights the importance of education in determining earning potential and suggests that investment in education could have long-term economic benefits. The study provides evidence for policy implications regarding educational incentives and workforce development strategies in Canada."
format: pdf
toc: true
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(here)
library(ggplot2)
library(kableExtra)
library(modelsummary)
library(rstanarm)
```

# Introduction {#sec-intro}

The relationship between education and earnings is a well-established topic of interest within the field of labor economics, providing insights into the broader socio-economic fabric of a country. As Canada's economy continues to diversify and specialize in various sectors, the value of education in this landscape remains a critical question for policymakers and the public. While previous studies have explored this link, there is an ongoing need to update and deepen our understanding of how this dynamic plays out in the modern economy.

This paper narrows the focus to the Canadian labor market, where the interplay between educational achievement and wages within the core working age group, those aged 25-54, offers a mirror to the efficacy and value of educational advancements in the workplace. This demographic is selected for its relevance to the labor force and policy implications, excluding younger individuals still likely in the education system and older individuals, who may present outlier educational experiences that do not align with the central workforce.

This investigation seeks to address a gap in the current literature by providing a targeted analysis of the 25-54 age group in Canada since the year 2000, a period marked by rapid technological change and economic evolution. By applying a linear model to wage data categorized by education level, this study uncovers a positive correlation between educational attainment and hourly wages, suggesting that higher education can be linked to improved wage outcomes in this key demographic.

The structure of the paper is organized as follows: Following @sec-intro, @sec-data presents the data, detailing the data sources, analytical techniques, and the rationale behind the chosen methods. @sec-model then covers the specifics of the linear model analysis, laying out the statistical underpinnings that support our investigation. After that, @sec-result discusses the results, elaborating on the observed trends and patterns in wage rate data. @sec-discussion interprets these findings in light of the current economic and educational context in Canada, exploring potential factors influencing these trends, drawing connections to broader socio-economic issues, and providing suggestions for future research in this area.

## Estimand

In this study, our primary focus is to estimate the effect of education on hourly wages in Canada. Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level. By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums.

# Data {#sec-data}

This section aims to offer an insightful understanding of the dataset utilized in our analysis, which serves as the foundation for our examination of the relationship between education and average hourly wage rates in Canada, specifically within the 25-54 age demographic.

## Source and Methodology

This study uses a dataset sourced from the Open Government Portal of Canada [@citedata], specifically designed to track the correlation between educational attainment and hourly wages across various demographic segments of the Canadian workforce. The dataset covers the period from the year 2000 onwards, providing a longitudinal view of wage trends in relation to educational background.

Alternative datasets, such as those from Statistics Canada’s Labour Force Survey, were considered but were not selected due to their less detailed categorization of education levels and their broader focus on employment without specific wage breakdowns by education.

The data was processed and cleaned using `R` [@citeR], a powerful statistical programming language. The selection of this dataset was motivated by its direct relevance to the research question and its regular updates, ensuring that the data remains relevant for observing current trends. Initial data processing involved filtering the dataset to focus solely on entries categorized under 'Canada' in geography, eliminating data irrelevant to the national focus of this research. This step was crucial to maintaining the clarity and relevance of the analysis. For key operations, please refer to [Appendix -@sec-data-cleaning].

## Variables

To better understand the data, key variables extracted for this study include the year of data collection, education level, age group, and the average hourly wage rate. Notably, we have streamlined the 'Education level' from a broad range of categories into a numeric variable that aligns with ascending educational attainment, facilitating a quantitative analysis of its impact on wages. The age group has been limited to the 25-54 years range to focus on the most economically active segment of the population, avoiding potential outliers from younger individuals with less work experience and older individuals whose advanced education, such as doctoral degrees, could distort the analysis. Wage rates are presented in Canadian dollars per hour, reflecting pre-tax earnings, inclusive of tips, commissions, and bonuses.

```{r}
#| label: tbl-display-clean-data
#| tbl-cap: First Ten Rows of Hourly Wage Across 2000-2019 After Data Processing
#| echo: false

cleaned_data <- read.csv(here::here("data/analysis_data/analysis_data.csv"))

cleaned_data |>
  select(Year, Education.level, Age.group, Avg.hourly.wage.rate) |>
  slice(1:10) |>
  kable(
    col.names = c("Year", "Education Level", "Age", "Hourly Wage"),
    digits = 1,
    booktabs = TRUE
  )
```
@tbl-display-clean-data, created with `kableExtra` [@citekableExtra], outlines the first ten records of our dataset, featuring average hourly wages for Canadians aged 25-54 across varying levels of education from the years 2000 to 2019. This snapshot reveals the variable structure of our analysis, which includes the `Year` of wage data, the defined `Education level` ranging from "0 - 8 years" to "Above bachelor’s degree," and the consistent `Age group` focus. It provides an early indication of the ascending trend in wages with higher educational attainment, setting the stage for our deeper investigation into the economic value of education within Canada's workforce.

To transition from a tabular overview to a more graphical representation, we will visualize the data, which will allow us to spot patterns and trends that are not immediately apparent in numerical form.

```{r fig.width=7, fig.height=3}
#| label: fig-display-clean-data
#| fig-cap: Average Hourly Wages by Educational Attainment, 2000-2019
#| echo: false

ggplot(cleaned_data, aes(x = Year, y = Avg.hourly.wage.rate, color = Education.level)) +
  geom_point(alpha = 0.7, size = 2) + 
  labs(
    x = "Year",
    y = "Average Hourly Wage Rate (CAD)",
    color = "Education Level"
  ) +
  theme_minimal(base_size = 10) + 
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8),
    axis.text.y = element_text(size = 8),
    legend.position = "right",
    legend.text = element_text(size = 8)
  ) +
  scale_color_brewer(palette = "Set1")  # Keep the color palette for distinction
```
@fig-display-clean-data, built by `ggplot2` [@citeggplot], illustrates the distribution of average hourly wage rates across different educational levels from 2000 to 2019 for Canadian workers aged 25-54. Each dot represents the wage rate for a specific education level in a given year, with color coding to distinguish between educational categories, ranging from "0 – 8 years" to "University degree." The spread and upward trend of the points suggest that individuals with higher educational qualifications tend to have higher average hourly wages, and this trend persists over the 20-year period.

## Measurements

The measurement of variables within this dataset was handled to ensure accuracy and relevance. The 'Average Hourly Wage Rate' is calculated based on actual wages reported by employers, providing an objective measure of income. This data is then vetted and standardized by the Government of Ontario to ensure it reflects accurate and fair representations of wages across different demographics.

For educational attainment, data entry occurs as individuals enter the workforce or update their qualifications, with categorizations reflecting the highest level of formal education completed. This data is collected through surveys and employment records, often as part of broader demographic data collection efforts by governmental agencies.

Correlations between the variables will be examined in @sec-model, providing insights into how education levels are associated with hourly wages. By understanding these relationships, we can better infer the potential impact of education on earnings within the Canadian economy.

# Model {#sec-model}

The objective of our modeling approach is to quantify the relationship between educational attainment and hourly wages. Our analysis employs a Bayesian framework to assess how changes in the level of education correlate with variations in wage rates among Canadian workers aged 25-54.

In our analysis, we utilized a simple linear regression model, a technique well-suited for examining the relationship between a continuous dependent variable and one or more independent variables. Given that our dependent variable, `average hourly wage rate`, is continuous, a normal or Gaussian linear regression model is appropriate. This model assumes that the distribution of the wage rate, given the level of education, follows a normal distribution. This assumption allows for a straightforward interpretation of the parameters and is commonly used in practice. The use of a normal distribution for the error term is a standard choice, as it facilitates the estimation of the model parameters using maximum likelihood estimation. The simplicity and interpretability of this model, coupled with its appropriateness for the data at hand, make it a good fit for our study's objectives.

Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ represent the average hourly wage rate for the $i^{th}$ individual. The predictor variable, $x_i$, corresponds to the numeric value assigned to each education level. The model can be described by the following equations:

$$
\begin{aligned}
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \beta_0 +\beta_1x_i\\
\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1) \\
\end{aligned}
$$

The model is executed in `R` [@citeR] using the `rstanarm` package [@citerstanarm]. Default priors from `rstanarm` [@citerstanarm] are used, set with a mean of zero and a conservative standard deviation.

## Model justification

Given the economic theory and previous findings, we anticipate that higher educational qualifications would correspond to higher hourly wages. This is encapsulated in the assumption that $\beta_i$ should have a positive effect on $y_i$. Through Bayesian analysis, we aim to capture the degree to which educational attainment can predict wage rates, thus providing evidence to support or refine this hypothesis.

We selected a Gaussian linear regression model to examine the relationship between education level and average hourly wages because our dependent variable, the wage rate, is a continuous data point that typically follows a normal distribution. Linear regression is a straightforward method for assessing the impact of one or more independent variables on a dependent variable. In this case, it helps us understand how increasing education levels, which we have numerically encoded, relate to wage rates.

Further justification for this model comes from its foundation in the central limit theorem, which suggests that with a large sample size, our wage rate data should be normally distributed. Moreover, the expected positive correlation between education and wages aligns with existing economic studies, reinforcing our theoretical framework. The simplicity of the model aids in avoiding overfitting, ensuring that the results are generalizable and relevant to a broader population. The regression model's assumptions—linearity, independence, and normality of residuals—are all conditions that our dataset reasonably meets, making a Gaussian linear regression not only a convenient choice but also statistically appropriate for our analysis.

# Results {#sec-result}

@sec-result explores the relationship between education and average hourly wages in Canada. Utilizing a dataset that captures 20 years of wage data across varying educational levels, we apply a linear modeling approach to uncover patterns and draw conclusions about the economic value of educational advancement. Below, we present the results of our model and its implications.

```{r}
#| label: tbl-summary-model
#| tbl-cap: Summary of the Linear Regression Model
#| echo: false
#| warning: false

linear_model <- readRDS(file = here::here("models/linear_model.rds"))

modelsummary(
  list(
    "Linear Model" = linear_model
  )
)
```

The results obtained from our linear model, which predicts the average hourly wage rate based on the numeric representation of education levels, show a clear and positive relationship between educational attainment and wage rate. @tbl-summary-model indicates that the intercept is estimated at 16.143, suggesting that when the education level is at the baseline (0 - not accounted for in the numeric scale), the average hourly wage rate would be approximately 16.14. Importantly, the coefficient for Education_numeric is positive (1.672), supporting the hypothesis that higher educational levels correlate with higher wage rates. The model used 180 observations for the analysis, with an R-squared value of 0.610, which means that approximately 61% of the variance in the average hourly wage rate is explained by the model. The adjusted R-squared, which accounts for the number of predictors in the model, is very close to the R-squared, indicating a good fit without overfitting. Notably, the RMSE (Root Mean Square Error) of 3.43 reflects the standard deviation of the prediction errors, which measures how spread out these errors are around the true regression line.

```{r fig.width=7, fig.height=4}
#| label: fig-visualize-model
#| fig-cap: Relationship between Hourly Wage Rate and Education Level
#| echo: false
#| warning: false

analysis_data <- read.csv(file = here::here("data/analysis_data/analysis_data.csv"))

plot1 <- analysis_data |>
  ggplot(aes(x = Education_numeric, y = Avg.hourly.wage.rate, colour = as.factor(Year))) +
  geom_point(alpha = 0.5) +
  geom_smooth(
    method = "lm",
    se = TRUE,
    color = "black",
    linetype = "dashed",
    formula = y ~ x
  ) +
  scale_colour_viridis_d() +  # Colorblind-friendly palette
  labs(
    x = "Education Level",  # Rename x-axis
    y = "AVG Hourly Wage Rate",  # Rename y-axis
    colour = "Year"  # Rename legend title
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
    axis.title.x = element_text(size = 12),  # Adjust x-axis title size if needed
    axis.title.y = element_text(size = 12),  # Adjust y-axis title size if needed
    legend.position = "right",
    legend.title = element_text(hjust = 0.5)
  ) +
  guides(colour = guide_legend(title = "Year", ncol = 2))
# Display the plot
print(plot1)
```
@fig-visualize-model illustrates the fitted linear regression line across the spread of average hourly wage rates for different education levels over the years 2000 to 2019. Each point in the scatter plot corresponds to the average hourly wage rate for a given education level, differentiated by color to represent each year. The dashed line across the plot signifies the best-fit linear regression line, which inclines upward as the education level increases. The diversity of colors along the regression line also reveals the variation in wage rates across different years, although the increasing trend is consistent throughout.

In conclusion, the results presented align with the initial hypothesis, showing that as education level increases, there is a corresponding increase in the average hourly wage rate. This validates the underlying assumption that more advanced education tends to lead to better-paying job opportunities. The analysis effectively captures the essence of the data, allowing for a meaningful interpretation that leads us into the discussion. In the next @sec-discussion, we will discuss the implications of these findings, examining their relevance to current labor market theories and the potential for informing education and economic policy decisions.

# Discussion {#sec-discussion}

In this paper, we investigated the relationship between educational attainment and average hourly wage rates from 2000 to 2019. By deploying a linear Bayesian model, we have been able to highlight the correlation between higher educational attainment and increased wage rates, aligning with our original hypothesis.

## Findings

Our analysis reveals a clear, positive correlation between education level and average hourly wage rates across the data spanning from 2000 to 2019. The consistent upward trend in wage rates with higher education levels across the years underscores the value of educational attainment in the labor market. Despite some fluctuations, the general pattern suggests that investments in education are likely to be economically beneficial for individuals in the age group of 25-54 years.

## Insights on Education and Wages

The results of this study offer clear evidence that educational attainment is closely tied to wage earnings, highlighting the premium that the labor market places on higher qualifications. This finding is consistent with current labor market theories that emphasize the importance of human capital—suggesting that investments in education are indeed reflected in higher wages. Specifically, our data supports the view that a well-educated workforce is critical for economic growth and competitiveness in today's globalized world, which increasingly values knowledge and specialized skills.

From a policy perspective, these insights could serve as a catalyst for reform, underscoring the need for investments in education. Policymakers might use this information to support sectors that are likely to offer greater economic returns for educational investments. This could involve bolstering STEM (Science, Technology, Engineering, and Mathematics) education, for example, or providing incentives for pursuing in-demand fields such as healthcare or information technology.

In conclusion, this study's findings highlight the significance of education as a determinant of economic success and reinforce the importance of informed educational and economic policymaking. For policymakers, the imperative is clear: to design and implement educational strategies that not only increase access to education but also ensure that the education provided is relevant and responsive to the current and future needs of the labor market.

## Limitations and Future Research Directions

The primary limitation of this study is its relatively small sample size, with only 180 observations. This constraint may impact the generalization of the findings. Additionally, the focus on the age group of 25-54 years might not encapsulate the full spectrum of the workforce, especially considering the varied career stages and life circumstances of younger and older workers. Such a limited scope might overlook the effects of education on wages across different life stages.

Future research should aim to disaggregate the age groups to understand the differential impact of education on wages. Distinguishing between early-career individuals, mid-career professionals, and late-career individuals could yield more insights into the education-wage dynamic at different life stages. Additionally, expanding the sample size and including more varied age groups would enhance the representativeness and depth of the analysis.

\newpage

\appendix

# Appendix {-}

# Data Manipulation and Cleaning {#sec-data-cleaning}

During the data cleaning phase, the R packages `tidyverse` [@citetidyverse], `dplyr` [@citedplyr], and `arrow` [@citearrow] were utilized. The raw data was imported using `read_csv` from the `tidyverse` [@citetidyverse] package. Subsequent operations filtered the dataset to focus on Canadian data, both full-time and part-time. Additionally, the focus was placed on records that reported the average hourly wage rate, while entries with ambiguous educational levels such as "Some high school" and aggregate categories like "Total, all education levels" were excluded to maintain data clarity and relevance.

The analysis was specifically targeted at the demographic group aged between 25 and 54 years, as this range represents a prime working-age population. Columns important to the analysis were chosen and renamed for better readability and straightforward reference in subsequent analytical procedures. This process included renaming the columns for the year and average hourly wage rate to 'Year' and 'Avg hourly wage rate' respectively.

Education levels were organized in a logical order ranging from "0 - 8 years" of education to "Above bachelor's degree" using the `factor` function to convert them into a categorical variable with a specified level order. This categorization was further enhanced by creating an accompanying numeric variable that mapped these ordered education levels to integers, thus facilitating quantitative analysis.

The final step involved saving the cleaned and structured data. This was done using the `write_csv` function to generate a CSV file for broad compatibility and the `write_parquet` function from the `arrow` [@citearrow] package for a more compressed and efficient file format, both of which were stored in the `data/analysis_data` directory.

In all figures and tables, the library `here` [@citehere] was used to ensure that the file path should be accessible in all directories.

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1, we implement a posterior predictive check, which displays the overlap between the observed data (denoted by $y$) and the replicated data generated from the model (denoted by $y_{rep}$). The density of the replicated data largely coincides with the observed data, suggesting that our model does a reasonable job of capturing the underlying distribution of the observed average hourly wage rates. This alignment is crucial as it indicates that the model's predictions are consistent with the real-world data we aim to understand and that the model can be a reliable tool for further inference.

In @fig-ppcheckandposteriorvsprior-2, we compare the posterior distributions of our parameters with the priors. This comparison reveals how the observed data have updated our beliefs about the parameters. We see that the posterior distributions for the intercept and the education_numeric coefficient are more concentrated and shifted from the priors, indicating that the data provided substantial information to update our initial assumptions. The sigma parameter's posterior, which represents the standard deviation of the residuals, is also refined from its prior, showing that the data have informed us about the variability in wages that is not explained by education level alone.

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(linear_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(linear_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(linear_model, "trace")

plot(linear_model, "rhat")
```

In @fig-stanareyouokay-1, the trace plot demonstrates the sampling paths of the Markov chains for each parameter in the model. The trace for the intercept, education_numeric coefficient, and sigma parameter all display a ‘hairy caterpillar’ shape, indicating good mixing and convergence across multiple chains. This suggests that the model's parameter estimates are stable and reliable, as evidenced by the overlapping and intertwined paths, which is indicative of a well-fitting model.

@fig-stanareyouokay-2, the Rhat plot, presents the R-hat statistics for each parameter, which is a measure of convergence. We see that all values at or very close to 1.00. This is a desirable outcome, implying that further sampling would likely yield little additional information and that the posterior estimates can be considered reliable for inference.

\newpage

# References

### 2024-11-26 scripts/00-install_packages.R

@@ -1,0 +1,19 @@
+#### Preamble ####
+# Purpose: Install necessary packages
+# Author: Yuanyi (Leo) Liu
+# Date: 31 March 2024
+# Contact: leoy.liu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: -
+# Any other information needed? -
+
+
+#### Install packages ####
+install.packages("tidyverse")
+install.packages("arrow")
+install.packages("dplyr")
+install.packages("rstanarm")
+install.packages("here")
+install.packages("ggplot2")
+install.packages("kableExtra")
+install.packages("modelsummary")

### 2024-11-26 scripts/01-simulate_data.R

@@ -1,0 +1,60 @@
+#### Preamble ####
+# Purpose: Simulates data
+# Author: Yuanyi (Leo) Liu
+# Email: leoy.liu@mail.utoronto.ca
+# Date: 31 March 2024
+# License: MIT
+# Pre-requisites: -
+# Any other information needed? -
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(dplyr)
+
+
+#### Simulate data ####
+set.seed(853)
+
+# Set parameters for simulation
+years <- 2000:2019
+education_levels <- c("0 - 8 years", "High school graduate", 
+                      "Post-secondary certificate or diploma", "Trade certificate or diploma", 
+                      "Community college, CEGEP", "University certificate below bachelors degree", 
+                      "University degree", "Bachelor's degree", "Above bachelor's degree")
+age_group <- "25-54 years"
+
+# Simulate data
+simulated_data <- expand.grid(Year = years, 
+                              Education_level = education_levels, 
+                              Age_group = age_group)
+
+# Assign education_numeric based on order in education_levels
+simulated_data$Education_numeric <- as.integer(factor(simulated_data$Education_level, levels = education_levels))
+
+# Define the wage rate ranges for each education level
+wage_rate_ranges <- list(
+  `0 - 8 years` = c(11, 13),
+  `High school graduate` = c(12, 14),
+  `Post-secondary certificate or diploma` = c(13, 15),
+  `Trade certificate or diploma` = c(14, 16),  
+  `Community college, CEGEP` = c(16, 18),
+  `University certificate below bachelors degree` = c(18, 20),
+  `University degree` = c(21, 24),
+  `Bachelor's degree` = c(24, 27),
+  `Above bachelor's degree` = c(28, 32))
+
+# Apply the wage rate ranges to the Avg_hourly_wage_rate column based on Education_level
+simulated_data <- simulated_data %>%
+  rowwise() %>%
+  mutate(Avg_hourly_wage_rate = round(runif(1, min = wage_rate_ranges[[Education_level]][1], 
+                                            max = wage_rate_ranges[[Education_level]][2]), 2))
+
+# Reset the order of the rows and remove the row names
+simulated_data <- simulated_data %>%
+  ungroup() %>%
+  arrange(Year, Education_level) %>%
+  select(Year, Education_level, Age_group, Avg_hourly_wage_rate, Education_numeric)
+
+# View the first few rows of the simulated dataset
+head(simulated_data)

### 2024-11-26 scripts/02-data_cleaning.R

@@ -1,0 +1,46 @@
+#### Preamble ####
+# Purpose: Cleans the raw plane data
+# Author: Yuanyi (Leo) Liu
+# Date: 31 March 2024
+# Contact: leoy.liu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: raw_data.csv
+# Any other information needed? -
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(arrow)
+
+
+#### Clean data ####
+raw_data <- read_csv("data/raw_data/raw_data.csv")
+
+cleaned_data <-
+  raw_data |>
+  filter(Geography == "Canada",
+         `Type of work` == "Both full- and part-time",
+         Wages == "Average hourly wage rate",
+         `Education level` != "Some high school",
+         `Education level` != "Some post-secondary",
+         `Education level` != "Total, all education levels",
+         `Education level` != "PSE  (5,6,7,8,9))",
+         `Education level` != "No PSE  (0,1,2,3,4)",
+         YEAR >= 2000) |>
+  filter(`Age group` == "25-54 years") |>
+  select(YEAR, `Education level`, `Age group`, `Both Sexes`) |>
+  rename(Year = YEAR, `Avg hourly wage rate` = `Both Sexes`)
+
+education_levels_ordered <- c("0 - 8  years", "High school graduate", 
+                              "Post-secondary certificate or diploma", "Trade certificate or diploma", 
+                              "Community college, CEGEP", "University certificate below bachelors degree", 
+                              "University degree", "Bachelor's degree", "Above bachelor's degree")
+
+cleaned_data <- cleaned_data %>%
+  mutate(`Education level` = factor(`Education level`, levels = education_levels_ordered)) %>%
+  mutate(Education_numeric = as.numeric(`Education level`))
+
+
+#### Save data ####
+write_csv(cleaned_data, "data/analysis_data/analysis_data.csv")
+write_parquet(cleaned_data, "data/analysis_data/analysis_data.parquet")

### 2024-11-26 scripts/03-test_data.R

@@ -1,0 +1,47 @@
+#### Preamble ####
+# Purpose: Tests for dataset
+# Author: Yuanyi (Leo) Liu
+# Date: 31 March 2024
+# Contact: leoy.liu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: 02-data_cleaning.R
+# Any other information needed? -
+
+
+#### Workspace setup ####
+library(tidyverse)
+
+
+#### Test data ####
+# Load dataset
+analysis_data <- read.csv("data/analysis_data/analysis_data.csv", check.names = FALSE)
+
+# Test 1: Check for missing values in simulated_data
+test1_result <- !any(is.na(analysis_data))
+print(paste("Test 1 Passed:", test1_result))
+
+# Test 2: Check for duplicated rows in simulated_data
+test2_result <- !any(duplicated(analysis_data))
+print(paste("Test 2 Passed:", test2_result))
+
+# Test 3: Check that there are exactly 9 unique education levels
+expected_levels <- c("0 - 8  years", "High school graduate", 
+                     "Post-secondary certificate or diploma", "Trade certificate or diploma", 
+                     "Community college, CEGEP", "University certificate below bachelors degree", 
+                     "University degree", "Bachelor's degree", "Above bachelor's degree")
+
+actual_levels <- unique(analysis_data$`Education level`)
+test3_result <- length(actual_levels) == 9 && all(actual_levels %in% expected_levels)
+print(paste("Test 3 Passed:", test3_result))
+
+# Test 4: Check that Education_numeric matches the levels of Education_level
+test4_result <- all(analysis_data$Education_numeric == as.integer(factor(analysis_data$Education_level, levels = education_levels)))
+print(paste("Test 4 Passed:", test4_result))
+
+# Test 5: Check that all years are within the specified range (2000-2019)
+test5_result <- all(analysis_data$Year >= 2000 & analysis_data$Year <= 2019)
+print(paste("Test 5 Passed:", test5_result))
+
+# Test 6: Check that all age groups are "25-54 years"
+test6_result <- all(analysis_data$`Age group` == "25-54 years")
+print(paste("Test 6 Passed:", test6_result))

### 2024-11-26 scripts/04-model.R

@@ -1,0 +1,39 @@
+#### Preamble ####
+# Purpose: Models the dataset with simple linear regression
+# Author: Yuanyi (Leo) Liu
+# Date: 31 March 2024
+# Contact: leoy.liu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: 02-data_cleaning.R
+# Any other information needed? -
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(arrow)
+library(rstanarm)
+library(modelsummary)
+
+
+#### Read data ####
+analysis_data <- read.csv(file = "data/analysis_data/analysis_data.csv")
+
+
+### Model data ####
+first_model <-
+  stan_glm(
+    formula = Avg.hourly.wage.rate ~ Education_numeric,
+    data = analysis_data,
+    family = gaussian(),
+    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
+    prior_aux = exponential(rate = 1, autoscale = TRUE),
+    seed = 853
+  )
+
+
+#### Save model ####
+saveRDS(
+  first_model,
+  file = "models/linear_model.rds"
+)

### 2024-11-26 paper/paper.qmd
update scripts and readme

@@
 
+In this study, our primary focus is to estimate the effect of education on hourly wages in Canada.
+Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level.
+By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums.
+
 This investigation seeks to address a gap in the current literature by providing a targeted analysis of the 25-54 age group in Canada since the year 2000, a period marked by rapid technological change and economic evolution.
@@
 
-## Estimand
-
-In this study, our primary focus is to estimate the effect of education on hourly wages in Canada.
-Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level.
-By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums.
-
 # Data {#sec-data}
 
-This section aims to offer an insightful understanding of the dataset utilized in our analysis, which serves as the foundation for our examination of the relationship between education and average hourly wage rates in Canada, specifically within the 25-54 age demographic.
+This section aims to offer a understanding of the dataset utilized in our analysis, which serves as the foundation for our examination of the relationship between education and average hourly wage rates in Canada, specifically within the 25-54 age demographic.
 

### 2024-11-26 scripts/00-simulate_data.R

@@ -2,7 +2,7 @@
 # Purpose: Simulates data
 # Author: Yuanyi (Leo) Liu
 # Email: leoy.liu@mail.utoronto.ca
-# Date: 31 March 2024
+# Date: 26 November 2024
 # License: MIT
 # Pre-requisites: -
 # Any other information needed? -

### 2024-11-26 scripts/01-test_simulated_data.R

@@ -1,0 +1,94 @@
+#### Preamble ####
+# Purpose: Tests the structure and validity of the simulated "placeholder" dataset.
+# Author: Yuanyi (Leo) Liu
+# Date: 26 November 2024
+# Contact: leoy.liu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: 
+#   - The `tidyverse`, `testthat` packages must be installed and loaded
+#   - 00-simulate_data.R must have been run
+# Any other information needed? Make sure you are in the `Forecasting-the-2024-US-Presidential-Election` rproj
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(testthat)
+
+simulated_data <- read_csv("data/00-simulated_data/simulated_data.csv")
+
+# Test if the data was successfully loaded
+test_that("dataset was successfully loaded", {
+  expect_true(exists("simulated_data"))
+})
+
+
+#### Test data ####
+
+# Check if the dataset has 1000 rows
+test_that("dataset has 1000 rows", {
+  expect_equal(nrow(simulated_data), 1000)
+})
+
+# Check if the dataset has 11 columns
+test_that("dataset has 11 columns", {
+  expect_equal(ncol(simulated_data), 11)
+})
+
+# Check if the 'state' column contains only valid US state names
+valid_states <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware",
+                           "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
+                           "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
+                           "Missouri", "Montana", "Nebraska", "Nevada","New Hampshire", "New Jersey", "New Mexico", "New York",
+                           "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", 
+                           "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", 
+                           "West Virginia", "Wisconsin", "Wyoming", "District of Columbia", "National")
+
+test_that("'state' column contains only valid US state names", {
+  expect_true(all(simulated_data$state %in% valid_states))
+})
+
+# Check if there are any missing values in the dataset
+test_that("dataset contains no missing values", {
[…]

### 2024-11-26 scripts/02-clean_data.R

@@ -1,7 +1,7 @@
 #### Preamble ####
 # Purpose: Cleans the raw plane data
 # Author: Yuanyi (Leo) Liu
-# Date: 31 March 2024
+# Date: 26 November 2024
 # Contact: leoy.liu@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: raw_data.csv

### 2024-11-26 scripts/04-exploratory_data_analysis.R

@@ -1,0 +1,113 @@
+#### Preamble ####
+# Purpose: Exploratory data analyis on cleaned data.
+# Author: Yuanyi (Leo) Liu
+# Date: 26 November 2024
+# Contact: leoy.liu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: 
+#   - The `tidyverse`, `rstanarm` packages must be installed and loaded
+#   - 03-clean_data.R must have been run
+# Any other information needed? Make sure you are in the `Forecasting-the-2024-US-Presidential-Election` rproj
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(rstanarm)
+
+#### Read data ####
+analysis_data <- read_csv("data/02-analysis_data/analysis_data.csv")
+
+# General overview of the dataset
+glimpse(analysis_data)
+summary(analysis_data)
+
+
+#### Summary Statistics ####
+# Summary of numeric columns
+analysis_data %>%
+  select_if(is.numeric) %>%
+  summary()
+
+# Count unique values in categorical columns
+analysis_data %>%
+  select_if(is.character) %>%
+  summarise_all(~ n_distinct(.))
+
+#### Missing Values Analysis ####
+# Check for missing values in each column
+analysis_data %>%
+  summarise_all(~ sum(is.na(.)))
+
+
+#### Univariate Analysis ####
+# Distribution of numeric_grade
+ggplot(analysis_data, aes(x = numeric_grade)) +
+  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
+  labs(title = "Distribution of Poll Numeric Grades", x = "Numeric Grade", y = "Count")
+
+# Distribution of transparency_score
+ggplot(analysis_data, aes(x = transparency_score)) +
+  geom_histogram(bins = 20, fill = "lightgreen", color = "black") +
+  labs(title = "Distribution of Poll Transparency Scores", x = "Transparency Score", y = "Count")
+
+# Distribution of sample_size
+ggplot(analysis_data, aes(x = sample_size)) +
+  geom_histogram(bins = 30, fill = "coral", color = "black") +
+  labs(title = "Distribution of Poll Sample Sizes", x = "Sample Size", y = "Count") +
+  scale_x_log10()  # Log scale for better visualization
+
+# Distribution of percent (poll result percentage)
+ggplot(analysis_data, aes(x = pct)) +
+  geom_histogram(bins = 30, fill = "purple", color = "black") +
[…]

### 2024-11-26 paper/paper.qmd
update paper content

@@
 
-## Source and Methodology
+## Overview
 
@@
 
-## Variables
-
-To better understand the data, key variables extracted for this study include the year of data collection, education level, age group, and the average hourly wage rate.
-Notably, we have streamlined the 'Education level' from a broad range of categories into a numeric variable that aligns with ascending educational attainment, facilitating a quantitative analysis of its impact on wages.
-The age group has been limited to the 25-54 years range to focus on the most economically active segment of the population, avoiding potential outliers from younger individuals with less work experience and older individuals whose advanced education, such as doctoral degrees, could distort the analysis.
-Wage rates are presented in Canadian dollars per hour, reflecting pre-tax earnings, inclusive of tips, commissions, and bonuses.
-
-```{r}
-#| label: tbl-display-clean-data
-#| tbl-cap: First Ten Rows of Hourly Wage Across 2000-2019 After Data Processing
-#| echo: false
-
-cleaned_data <- read.csv(here::here("data/analysis_data/analysis_data.csv"))
-
-cleaned_data |>
-  select(Year, Education.level, Age.group, Avg.hourly.wage.rate) |>
-  slice(1:10) |>
-  kable(
-    col.names = c("Year", "Education Level", "Age", "Hourly Wage"),
-    digits = 1,
-    booktabs = TRUE
-  )
-```
-@tbl-display-clean-data, created with `kableExtra` [@citekableExtra], outlines the first ten records of our dataset, featuring average hourly wages for Canadians aged 25-54 across varying levels of education from the years 2000 to 2019.
-This snapshot reveals the variable structure of our analysis, which includes the `Year` of wage data, the defined `Education level` ranging from "0 - 8 years" to "Above bachelor’s degree," and the consistent `Age group` focus.
-It provides an early indication of the ascending trend in wages with higher educational attainment, setting the stage for our deeper investigation into the economic value of education within Canada's workforce.
-
-To transition from a tabular overview to a more graphical representation, we will visualize the data, which will allow us to spot patterns and trends that are not immediately apparent in numerical form.
-
-```{r fig.width=7, fig.height=3}
-#| label: fig-display-clean-data
-#| fig-cap: Average Hourly Wages by Educational Attainment, 2000-2019
-#| echo: false
-
-ggplot(cleaned_data, aes(x = Year, y = Avg.hourly.wage.rate, color = Education.level)) +
-  geom_point(alpha = 0.7, size = 2) + 
-  labs(
-    x = "Year",
-    y = "Average Hourly Wage Rate (CAD)",
-    color = "Education Level"
-  ) +
-  theme_minimal(base_size = 10) + 
-  theme(
-    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1, size = 8),
-    axis.text.y = element_text(size = 8),
-    legend.position = "right",
-    legend.text = element_text(size = 8)
-  ) +
-  scale_color_brewer(palette = "Set1")  # Keep the color palette for distinction
-```
-@fig-display-clean-data, built by `ggplot2` [@citeggplot], illustrates the distribution of average hourly wage rates across different educational levels from 2000 to 2019 for Canadian workers aged 25-54.
-Each dot represents the wage rate for a specific education level in a given year, with color coding to distinguish between educational categories, ranging from "0 – 8 years" to "University degree." The spread and upward trend of the points suggest that individuals with higher educational qualifications tend to have higher average hourly wages, and this trend persists over the 20-year period.
-
-## Measurements
+## Measurement
 
@@
 By understanding these relationships, we can better infer the potential impact of education on earnings within the Canadian economy.
+
+## Outcome variables
+
+### Hourly Wages
+
+## Predictor variables
+
+### Education
+
+### Gender
+
+### Age
 
@@
 
+## Alternative model
+
 ## Model set-up
 
-Define $y_i$ represent the average hourly wage rate for the $i^{th}$ individual.
-The predictor variable, $x_i$, corresponds to the numeric value assigned to each education level.
-The model can be described by the following equations:
+The model predicts hourly wages using the following predictor variables:
+
+-
+
+The model takes the form:
 
@@
 
-The model is executed in `R` [@citeR] using the `rstanarm` package [@citerstanarm].
-Default priors from `rstanarm` [@citerstanarm] are used, set with a mean of zero and a conservative standard deviation.
+The model is executed in \texttt{R} [@citeR] using the \texttt{rstanarm} package [@rstanarm].
+Default priors from \texttt{rstanarm} [@rstanarm] are used, with the priors set to have a mean of zero and a moderate standard deviation to ensure a reasonable level of regularization.
 
@@
 
-Given the economic theory and previous findings, we anticipate that higher educational qualifications would correspond to higher hourly wages.
-This is encapsulated in the assumption that $\beta_i$ should have a positive effect on $y_i$.
-Through Bayesian analysis, we aim to capture the degree to which educational attainment can predict wage rates, thus providing evidence to support or refine this hypothesis.
-
-We selected a Gaussian linear regression model to examine the relationship between education level and average hourly wages because our dependent variable, the wage rate, is a continuous data point that typically follows a normal distribution.
-Linear regression is a straightforward method for assessing the impact of one or more independent variables on a dependent variable.
-In this case, it helps us understand how increasing education levels, which we have numerically encoded, relate to wage rates.
-
-Further justification for this model comes from its foundation in the central limit theorem, which suggests that with a large sample size, our wage rate data should be normally distributed.
-Moreover, the expected positive correlation between education and wages aligns with existing economic studies, reinforcing our theoretical framework.
-The simplicity of the model aids in avoiding overfitting, ensuring that the results are generalizable and relevant to a broader population.
-The regression model's assumptions—linearity, independence, and normality of residuals—are all conditions that our dataset reasonably meets, making a Gaussian linear regression not only a convenient choice but also statistically appropriate for our analysis.
-
 # Results {#sec-result}
-
-@sec-result explores the relationship between education and average hourly wages in Canada.
-Utilizing a dataset that captures 20 years of wage data across varying educational levels, we apply a linear modeling approach to uncover patterns and draw conclusions about the economic value of educational advancement.
-Below, we present the results of our model and its implications.
-
-```{r}
-#| label: tbl-summary-model
-#| tbl-cap: Summary of the Linear Regression Model
-#| echo: false
-#| warning: false
-
-linear_model <- readRDS(file = here::here("models/linear_model.rds"))
-
-modelsummary(
-  list(
-    "Linear Model" = linear_model
-  )
-)
-```
-
-The results obtained from our linear model, which predicts the average hourly wage rate based on the numeric representation of education levels, show a clear and positive relationship between educational attainment and wage rate.
-@tbl-summary-model indicates that the intercept is estimated at 16.143, suggesting that when the education level is at the baseline (0 - not accounted for in the numeric scale), the average hourly wage rate would be approximately 16.14.
-Importantly, the coefficient for Education_numeric is positive (1.672), supporting the hypothesis that higher educational levels correlate with higher wage rates.
-The model used 180 observations for the analysis, with an R-squared value of 0.610, which means that approximately 61% of the variance in the average hourly wage rate is explained by the model.
-The adjusted R-squared, which accounts for the number of predictors in the model, is very close to the R-squared, indicating a good fit without overfitting.
-Notably, the RMSE (Root Mean Square Error) of 3.43 reflects the standard deviation of the prediction errors, which measures how spread out these errors are around the true regression line.
-
-```{r fig.width=7, fig.height=4}
-#| label: fig-visualize-model
-#| fig-cap: Relationship between Hourly Wage Rate and Education Level
-#| echo: false
-#| warning: false
-
-analysis_data <- read.csv(file = here::here("data/analysis_data/analysis_data.csv"))
-
-plot1 <- analysis_data |>
-  ggplot(aes(x = Education_numeric, y = Avg.hourly.wage.rate, colour = as.factor(Year))) +
-  geom_point(alpha = 0.5) +
-  geom_smooth(
-    method = "lm",
-    se = TRUE,
-    color = "black",
-    linetype = "dashed",
-    formula = y ~ x
-  ) +
-  scale_colour_viridis_d() +  # Colorblind-friendly palette
-  labs(
-    x = "Education Level",  # Rename x-axis
-    y = "AVG Hourly Wage Rate",  # Rename y-axis
-    colour = "Year"  # Rename legend title
-  ) +
-  theme_classic() +
-  theme(
-    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
-    axis.title.x = element_text(size = 12),  # Adjust x-axis title size if needed
-    axis.title.y = element_text(size = 12),  # Adjust y-axis title size if needed
-    legend.position = "right",
-    legend.title = element_text(hjust = 0.5)
-  ) +
-  guides(colour = guide_legend(title = "Year", ncol = 2))
-# Display the plot
-print(plot1)
-```
-@fig-visualize-model illustrates the fitted linear regression line across the spread of average hourly wage rates for different education levels over the years 2000 to 2019.
-Each point in the scatter plot corresponds to the average hourly wage rate for a given education level, differentiated by color to represent each year.
-The dashed line across the plot signifies the best-fit linear regression line, which inclines upward as the education level increases.
-The diversity of colors along the regression line also reveals the variation in wage rates across different years, although the increasing trend is consistent throughout.
-
-In conclusion, the results presented align with the initial hypothesis, showing that as education level increases, there is a corresponding increase in the average hourly wage rate.
-This validates the underlying assumption that more advanced education tends to lead to better-paying job opportunities.
-The analysis effectively captures the essence of the data, allowing for a meaningful interpretation that leads us into the discussion.
-In the next @sec-discussion, we will discuss the implications of these findings, examining their relevance to current labor market theories and the potential for informing education and economic policy decisions.
 
@@
 
-In this paper, we investigated the relationship between educational attainment and average hourly wage rates from 2000 to 2019.
-By deploying a linear Bayesian model, we have been able to highlight the correlation between higher educational attainment and increased wage rates, aligning with our original hypothesis.
-
 ## Findings
-
-Our analysis reveals a clear, positive correlation between education level and average hourly wage rates across the data spanning from 2000 to 2019.
-The consistent upward trend in wage rates with higher education levels across the years underscores the value of educational attainment in the labor market.
-Despite some fluctuations, the general pattern suggests that investments in education are likely to be economically beneficial for individuals in the age group of 25-54 years.
 
@@
 
-The results of this study offer clear evidence that educational attainment is closely tied to wage earnings, highlighting the premium that the labor market places on higher qualifications.
-This finding is consistent with current labor market theories that emphasize the importance of human capital—suggesting that investments in education are indeed reflected in higher wages.
-Specifically, our data supports the view that a well-educated workforce is critical for economic growth and competitiveness in today's globalized world, which increasingly values knowledge and specialized skills.
-
-From a policy perspective, these insights could serve as a catalyst for reform, underscoring the need for investments in education.
-Policymakers might use this information to support sectors that are likely to offer greater economic returns for educational investments.
-This could involve bolstering STEM (Science, Technology, Engineering, and Mathematics) education, for example, or providing incentives for pursuing in-demand fields such as healthcare or information technology.
-
-In conclusion, this study's findings highlight the significance of education as a determinant of economic success and reinforce the importance of informed educational and economic policymaking.
-For policymakers, the imperative is clear: to design and implement educational strategies that not only increase access to education but also ensure that the education provided is relevant and responsive to the current and future needs of the labor market.
-
 ## Limitations and Future Research Directions
-
-The primary limitation of this study is its relatively small sample size, with only 180 observations.
-This constraint may impact the generalization of the findings.
-Additionally, the focus on the age group of 25-54 years might not encapsulate the full spectrum of the workforce, especially considering the varied career stages and life circumstances of younger and older workers.
-Such a limited scope might overlook the effects of education on wages across different life stages.
-
-Future research should aim to disaggregate the age groups to understand the differential impact of education on wages.
-Distinguishing between early-career individuals, mid-career professionals, and late-career individuals could yield more insights into the education-wage dynamic at different life stages.
-Additionally, expanding the sample size and including more varied age groups would enhance the representativeness and depth of the analysis.
 
@@
 # Appendix {-}
+
+# Census methodology overview and evaluation
+
+# Idealized methodology and survey
 
@@
 
-In @fig-ppcheckandposteriorvsprior-1, we implement a posterior predictive check, which displays the overlap between the observed data (denoted by $y$) and the replicated data generated from the model (denoted by $y_{rep}$).
-The density of the replicated data largely coincides with the observed data, suggesting that our model does a reasonable job of capturing the underlying distribution of the observed average hourly wage rates.
-This alignment is crucial as it indicates that the model's predictions are consistent with the real-world data we aim to understand and that the model can be a reliable tool for further inference.
-
-In @fig-ppcheckandposteriorvsprior-2, we compare the posterior distributions of our parameters with the priors.
-This comparison reveals how the observed data have updated our beliefs about the parameters.
-We see that the posterior distributions for the intercept and the education_numeric coefficient are more concentrated and shifted from the priors, indicating that the data provided substantial information to update our initial assumptions.
-The sigma parameter's posterior, which represents the standard deviation of the residuals, is also refined from its prior, showing that the data have informed us about the variability in wages that is not explained by education level alone.
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-pp_check(linear_model) +
-  theme_classic() +
-  theme(legend.position = "bottom")
-
-posterior_vs_prior(linear_model) +
-  theme_minimal() +
-  scale_color_brewer(palette = "Set1") +
-  theme(legend.position = "bottom") +
-  coord_flip()
-```
-
 ## Diagnostics
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-plot(linear_model, "trace")
-
-plot(linear_model, "rhat")
-```
-
-In @fig-stanareyouokay-1, the trace plot demonstrates the sampling paths of the Markov chains for each parameter in the model.
-The trace for the intercept, education_numeric coefficient, and sigma parameter all display a ‘hairy caterpillar’ shape, indicating good mixing and convergence across multiple chains.
-This suggests that the model's parameter estimates are stable and reliable, as evidenced by the overlapping and intertwined paths, which is indicative of a well-fitting model.
-
-@fig-stanareyouokay-2, the Rhat plot, presents the R-hat statistics for each parameter, which is a measure of convergence.
-We see that all values at or very close to 1.00.
-This is a desirable outcome, implying that further sampling would likely yield little additional information and that the posterior estimates can be considered reliable for inference.
 

### 2024-12-02 paper/paper.qmd
update scripts, intro and data

@@
 title: "Determinants of Wage Variation in Canada"
-subtitle: "The Influence of Education, Gender, and Age on Canadian Earnings"
+subtitle: "Older Workers, Men, and Higher Education Earn Higher Average Hourly Wages"
 author: 
   - Yuanyi (Leo) Liu
-thanks: "Code and data are available at: [Determinants of Wage Variation in Canada](https://github.com/leoyliu/Determinants-of-Wage-Variation-in-Canada)"
+thanks: "Code and data are available at: [Determinants of Wage Variation in Canada](https://github.com/leoyliu/Determinants-of-Wage-Variation-in-Canada)."
 date: today
 date-format: long
-abstract: "In this paper, we explore the connection between education and hourly wages in Canada, focusing on data from the year 2000.
-Our findings indicate a clear trend: higher educational levels correlate with increased average hourly wages for individuals aged 25 to 54.
-This research highlights the importance of education in determining earning potential and suggests that investment in education could have long-term economic benefits.
-The study provides evidence for policy implications regarding educational incentives and workforce development strategies in Canada."
+abstract: "This paper examines wage disparities in Canada by analyzing the effects of education, gender, and age on average hourly earnings.
+The findings indicate that older workers consistently earn higher wages, men earn more than women across all groups, and higher levels of education are strongly associated with increased wages.
+These patterns highlight systemic differences in earnings tied to demographic and educational factors.
+Understanding these disparities provides important context for policymakers aiming to address wage inequality and improve economic outcomes."
 format: pdf
@@
 
-The relationship between education and earnings is a well-established topic of interest within the field of labor economics, providing insights into the broader socio-economic fabric of a country.
-As Canada's economy continues to diversify and specialize in various sectors, the value of education in this landscape remains a critical question for policymakers and the public.
-While previous studies have explored this link, there is an ongoing need to update and deepen our understanding of how this dynamic plays out in the modern economy.
+Wages are a fundamental component of economic well-being, influencing individual livelihoods, workforce productivity, and broader societal equity.
+Understanding the factors that shape earnings is essential for addressing wage inequality and improving economic opportunities.
+In Canada, wage disparities persist across demographic and educational groups, raising important questions about the relationship between age, gender, and education and their roles in determining hourly wages.
 
-This paper narrows the focus to the Canadian labor market, where the interplay between educational achievement and wages within the core working age group, those aged 25-54, offers a mirror to the efficacy and value of educational advancements in the workplace.
-This demographic is selected for its relevance to the labor force and policy implications, excluding younger individuals still likely in the education system and older individuals, who may present outlier educational experiences that do not align with the central workforce.
+This paper analyzes how average hourly wages in Canada vary based on three key factors: age, gender, and education level.
+Using data spanning two decades, we employ a linear regression model to quantify the relationships between these factors and earnings.
+The analysis examines whether older workers earn higher wages, how gender influences pay, and the extent to which education impacts earnings.
+These questions are important in the ongoing concerns about income inequality and the need for policies that promote equitable pay structures.
+While previous studies have documented broad patterns of wage inequality, this paper offers a focused examination of specific demographic and educational factors, providing a deeper understanding of their combined effects.
 
-In this study, our primary focus is to estimate the effect of education on hourly wages in Canada.
-Our estimand is the incremental average hourly wage rate for individuals within the specified education levels, compared to the next lower education level.
-By quantifying this effect, we aim to capture the economic value of educational attainment and how it translates into wage premiums.
+In this study, our estimand is the average hourly wage rate in Canada.
+The object of the estimation is the average hourly wage rate across different demographic and educational groups based on the data.
+By modeling these relationships, the study quantifies the extent to which these factors contribute to observed wage disparities, providing a detailed understanding of their relative influence within the Canadian labor market.
 
-This investigation seeks to address a gap in the current literature by providing a targeted analysis of the 25-54 age group in Canada since the year 2000, a period marked by rapid technological change and economic evolution.
-By applying a linear model to wage data categorized by education level, this study uncovers a positive correlation between educational attainment and hourly wages, suggesting that higher education can be linked to improved wage outcomes in this key demographic.
+The findings demonstrate that wages increase with age, particularly for workers over 55, while men consistently earn more than women across all groups.
+Higher levels of education are strongly associated with higher wages, with university graduates earning substantially more than those with lower educational attainment.
+These results highlight persistent disparities that reflect both structural and systemic influences on earnings.
+By examining these relationships, this paper provides a basis for understanding wage inequality in Canada and informs discussions about potential policy interventions.
 
-The structure of the paper is organized as follows: Following @sec-intro, @sec-data presents the data, detailing the data sources, analytical techniques, and the rationale behind the chosen methods.
-@sec-model then covers the specifics of the linear model analysis, laying out the statistical underpinnings that support our investigation.
-After that, @sec-result discusses the results, elaborating on the observed trends and patterns in wage rate data.
-@sec-discussion interprets these findings in light of the current economic and educational context in Canada, exploring potential factors influencing these trends, drawing connections to broader socio-economic issues, and providing suggestions for future research in this area.
+The structure of the paper is organized as follows: following @sec-intro, @sec-data presents the data collection and cleaning process, along with an overview of the variables used in the analysis.
+@sec-model explains the chosen model and why it is appropriate for modeling average hourly wages.
+Then, @sec-result provides the results, highlighting key trends and predictions.
+Eventually, @sec-discussion concludes with a discussion of the findings, addressing wage disparities, potential strategies to improve equity in Canadian earnings, and the limitations of the model used.
 
@@
 
-This section aims to offer a understanding of the dataset utilized in our analysis, which serves as the foundation for our examination of the relationship between education and average hourly wage rates in Canada, specifically within the 25-54 age demographic.
-
 ## Overview
 
-This study uses a dataset sourced from the Open Government Portal of Canada [@citedata], specifically designed to track the correlation between educational attainment and hourly wages across various demographic segments of the Canadian workforce.
-The dataset covers the period from the year 2000 onwards, providing a longitudinal view of wage trends in relation to educational background.
+We use the statistical programming language R [@citeR] to process and analyze data on average hourly wages in Canada.
+The dataset [@citedata], published on January 11, 2020, covers data from January 1, 1997, to December 31, 2019, but for this analysis, we focus on data from 2000 onwards to ensure relevance to recent labor market trends.
+The dataset is maintained annually and provides detailed information on wages, demographics, and employment characteristics, making it well-suited for studying wage disparities in Canada.
+Following methodologies discussed in "Telling Stories with Data" [@tellingstories], we analyze wage patterns by aggregating data across multiple demographic groups to ensure a balanced and unbiased representation of wage disparities.
+For key operations, please refer to [Appendix -@sec-data-cleaning].
 
-Alternative datasets, such as those from Statistics Canada’s Labour Force Survey, were considered but were not selected due to their less detailed categorization of education levels and their broader focus on employment without specific wage breakdowns by education.
+The dataset includes variables such as age groups (15+, 25+, 25-34, 25-54, and 25-64), types of work (full-time and part-time), educational levels (e.g., 0-8 years, high school graduate, post-secondary certificate diploma, university degree), and immigration statuses (e.g., very recent immigrants, recent immigrants, established immigrants, non-landed immigrants, and Canadian-born individuals).
+Wages are provided as both average weekly and average hourly rates.
+For this analysis, we focus on the age groups "15-24 years", "25-54 years", and "55 years and over", the combined category of full- and part-time workers, and specific education levels, such as high school graduate, post-secondary certificate diploma, and university degree.
 
-The data was processed and cleaned using `R` [@citeR], a powerful statistical programming language.
-The selection of this dataset was motivated by its direct relevance to the research question and its regular updates, ensuring that the data remains relevant for observing current trends.
-Initial data processing involved filtering the dataset to focus solely on entries categorized under 'Canada' in geography, eliminating data irrelevant to the national focus of this research.
-This step was crucial to maintaining the clarity and relevance of the analysis.
-For key operations, please refer to [Appendix -@sec-data-cleaning].
+This filtering excludes ambiguous or aggregate categories such as "some high school," "some post-secondary," and "total landed immigrants" to ensure a clear and interpretable dataset.
+This filtering allows us to analyze relationships between wages and specific demographic factors while maintaining focus on education, gender, and age.
+Alternative datasets, such as those focusing on industry-specific wage trends or detailed time-series data, were not used because they do not provide the necessary demographic granularity.
+This dataset is uniquely positioned to address our research questions by capturing a broad view of wage disparities in Canada.
 
@@
 
-The measurement of variables within this dataset was handled to ensure accuracy and relevance.
-The 'Average Hourly Wage Rate' is calculated based on actual wages reported by employers, providing an objective measure of income.
-This data is then vetted and standardized by the Government of Ontario to ensure it reflects accurate and fair representations of wages across different demographics.
+Wage data is a measurement of economic outcomes based on information collected through the Labour Force Survey Special Tabulations conducted by Statistics Canada.
+Respondents' demographic information, including age, gender, education level, and employment type, is reported through standardized surveys administered to a representative sample of the Canadian workforce.
+Wage data, including both hourly and weekly rates, are derived from employer records and worker-reported earnings, ensuring broad coverage across different industries and occupations.
+More details on survey methodologies can be found in the [Labour Force Survey Special Tabulations](http://www5.statcan.gc.ca/cansim/a26?lang=eng&retrLang=eng&id=2820004&tabMode=dataTable&p1=-1&p2=9&srchLan=-1).
 
-For educational attainment, data entry occurs as individuals enter the workforce or update their qualifications, with categorizations reflecting the highest level of formal education completed.
-This data is collected through surveys and employment records, often as part of broader demographic data collection efforts by governmental agencies.
-
-Correlations between the variables will be examined in @sec-model, providing insights into how education levels are associated with hourly wages.
-By understanding these relationships, we can better infer the potential impact of education on earnings within the Canadian economy.
+Each row in the dataset represents an aggregated statistic for a specific combination of age group, education level, gender, and type of work.
+For example, the average hourly wage rate for "25-54 years" old male workers with a "university degree" is calculated by pooling the wages of individuals who fit this description, smoothing out individual variations to provide a general trend for this subgroup.
+By categorizing variables such as education levels into distinct groups (e.g., "high school graduate," "post-secondary certificate diploma"), the dataset allows for consistent comparisons across demographic segments.
 
@@
 
+The primary outcome variable in this study is the `Average_hourly_wages`, measured in Canadian dollars.
+This variable reflects the mean earnings of individuals across various demographic and employment categories.
+The hourly wage data are derived from the Labour Force Survey, which collects self-reported income information and supplements it with employer records where available.
+These averages are calculated by dividing reported weekly wages by the number of hours worked, standardizing earnings across full-time and part-time workers.
+
+```{r}
+#| label: tbl-display-clean-data
+#| tbl-cap: "Summary Statistics for Average Hourly Wages in Canada (2000–2019), Including Minimum, Maximum, Mean, Median, and Standard Deviation"
+#| echo: false
+
+cleaned_data <- read.csv(here::here("data/02-analysis_data/analysis_data.csv"))
+
+hourly_wage_summary <- cleaned_data %>%
+  summarise(
+    Min_Wage = min(Average_hourly_wages, na.rm = TRUE),
+    Max_Wage = max(Average_hourly_wages, na.rm = TRUE),
+    Mean_Wage = mean(Average_hourly_wages, na.rm = TRUE),
+    Median_Wage = median(Average_hourly_wages, na.rm = TRUE),
+    SD_Wage = sd(Average_hourly_wages, na.rm = TRUE)
+  )
+
+hourly_wage_summary %>%
+  kable(
+    col.names = c("Minimum Wage", "Maximum Wage", "Mean Wage", "Median Wage", "Standard Deviation"),
+    digits = 2,
+    booktabs = TRUE
+  )
+```
+
+@tbl-display-clean-data provides an overview of the distribution of average hourly wages in Canada from 2000 to 2019.
+The wages range from a minimum of \$7.60 to a maximum of \$43.39, with a mean of \$21.48 and a median of \$20.48, indicating that most wages are concentrated around the middle of the distribution, while a smaller proportion of higher earners pushes the maximum upward.
+The standard deviation of $7.40 reflects notable variability, emphasizing differences in earning potential influenced by factors such as education, gender, and age.
+
+```{r}
+#| label: fig-avg-wage
+#| fig-cap: "Yearly Trend of Average Hourly Wages in Canada (2000–2019): Steady Growth Reflecting Changes in the Labor Market."
+#| echo: false
+#| warning: false
+
+cleaned_data %>%
+  group_by(Year) %>%
+  summarise(Average_Wage = round(mean(Average_hourly_wages, na.rm = TRUE), 2)) %>%
+  ggplot(aes(x = Year, y = Average_Wage)) +
+  geom_line(color = "darkblue", size = 1) +
+  geom_point(color = "lightblue", size = 1.5) +
+  geom_label(
+    aes(label = Average_Wage),
+     nudge_x = 0.25,
+    nudge_y = 0.25,
+    check_overlap = TRUE,
+    size = 1.5
+  ) +
+  labs(
+    x = "Year",
+    y = "Average Hourly Wage",
+  ) +
+  theme_minimal()
+```
+
+@fig-avg-wage illustrates the trend of average hourly wages in Canada from 2000 to 2019.
+Each dot represents the average wage for a specific year, and the overall shows a upward trend.
+The consistent increase in wages suggests improvements in earnings across the labor market, potentially driven by inflation, economic growth, and changes in workforce composition.
+
 ## Predictor variables
@@
 
+The `Education_level` variable represents individuals' highest level of formal education attained and is categorized into levels ranging from "0-8 years" to "Above bachelor's degree." This variable reflects the influence of educational attainment on earning potential.
+The data for education levels are self-reported in the Labour Force Survey, capturing a wide range of qualifications, including high school completion, trade certifications, and university degrees.
+
+```{r fig.width=5, fig.height=5}
+#| label: fig-education-level
+#| fig-cap: "Average hourly wages in Canada by education level from 2000 to 2019.
+Each box represents the wage distribution for a specific education level, with higher education levels corresponding to higher median wages."
+#| echo: false
+#| warning: false
+
+cleaned_data %>%
+  ggplot(aes(
+    x = reorder(Education_level, Average_hourly_wages, median),
+    y = Average_hourly_wages,
+    fill = reorder(Education_level, Average_hourly_wages, median)
+  )) +
+  geom_boxplot(color = "black", outlier.color = "red", outlier.size = 1.5) +
+  scale_fill_manual(
+    values = colorRampPalette(c("mistyrose", "darkred"))(length(unique(cleaned_data$Education_level)))
+  ) +
+  labs(
+    x = "Education Level",
+    y = "Average Hourly Wage",
+    fill = "Education Level",
+  ) +
+  theme_minimal() +
+  theme(
+    legend.position = "none",
+    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
+    axis.text.y = element_text(size = 8)
+  )
+```
+
+@fig-education-level presents the distribution of average hourly wages in Canada from 2000 to 2019 across different education levels.
+Higher education levels, such as "University degree" and "Above bachelor's degree," are associated with higher median wages, while lower education levels, like "0–8 years," show significantly lower wages.
+The variability within each education level, shown by the spread of the boxes and whiskers, indicates differences in earning potential within these groups.
+For instance, individuals with an "Above bachelor's degree" earn higher wages on average and exhibit greater wage variability compared to those with "0–8 years" of education.
+
 ### Gender
 
+The `Gender` variable indicates whether an individual identifies as "Male" or "Female".
+The data are reported as binary classification for simplicity.
+This variable allows for an investigation of the persistent gender wage gap, capturing differences in average hourly wages for men and women.
+Including gender in the analysis is essential for highlighting inequalities in the labor market and understanding how these disparities persist across educational levels and age groups.
+
+```{r}
+#| label: fig-gender
+#| fig-cap: "Average hourly wages in Canada by gender from 2000 to 2019.
+Each violin plot represents the wage distribution for males and females, with males showing higher median wages and a broader range of earnings compared to females."
+#| echo: false
+#| warning: false
+
+cleaned_data %>%
+  ggplot(aes(x = Gender, y = Average_hourly_wages, fill = Gender)) +
+  geom_violin(trim = FALSE, alpha = 0.6) +
+  geom_jitter(width = 0.2, size = 1, alpha = 0.7, color = "black") +
+  labs(
+    x = "Gender",
+    y = "Average Hourly Wage",
+  ) +
+  theme_minimal()
+```
+
+@fig-gender compares the distribution of average hourly wages in Canada from 2000 to 2019 by gender.
+It displays the spread of wages for each gender, with males generally earning higher wages as shown by the broader and higher distribution.
+The jittered points represent individual wage observations, further emphasizing the overlap and differences in wage ranges.
+While both distributions have a similar shape, the median wage for males is noticeably higher than that for females, reflecting a persistent gender wage gap over the period analyzed.
+
 ### Age
+
+The `Age_group` variable divides individuals into three categories: "15-24 years", "25-54 years", and "55 years and over".
+These groups represent different career stages, from early employment to peak earning years and late career.
+The Labour Force Survey provides these classifications to standardize analyses of how wages vary with experience and age.
+Younger workers generally report lower wages due to limited experience, while mid-career and older workers often earn higher wages, reflecting accumulated skills and knowledge over time.
+
+```{r}
+#| label: fig-age
+#| fig-cap: "Average hourly wages in Canada by age group from 2000 to 2019.
+Each curve represents the wage distribution for a specific age group, with older age groups showing higher average wages and broader distributions compared to younger groups."
+#| echo: false
+#| warning: false
+
+cleaned_data %>%
+  ggplot(aes(x = Average_hourly_wages, fill = Age_group)) +
+  geom_density(alpha = 0.7, color = "black", size = 0.3) + 
+  labs(
+    x = "Average Hourly Wage",
+    y = "Density",
+  ) +
+  scale_fill_manual(
+    values = c("15-24 years" = "lavender", 
+               "25-54 years" = "mediumorchid", 
+               "55 years and over" = "purple4")
+  ) + 
+  theme_minimal()
+
+```
+
+@fig-age illustrates the distribution of average hourly wages in Canada from 2000 to 2019 across three age groups: "15-24 years," "25-54 years," and "55 years and over." The youngest group, "15-24 years," exhibits the lowest wages, with a narrow distribution concentrated at the lower end.
+In contrast, the "25-54 years" and "55 years and over" groups show higher wages and broader distributions, reflecting greater variability in earnings as individuals gain experience and qualifications.
+The older age group generally earns the highest wages, highlighting the strong association between age, experience, and earning potential.
 

### 2024-12-02 scripts/00-simulate_data.R

@@ -1,20 +1,19 @@
 #### Preamble ####
-# Purpose: Simulates data
+# Purpose: Simulates a dataset of Canadian average hourly wages from 2000 to 2019 to explore what values might appear in the dataset.
 # Author: Yuanyi (Leo) Liu
-# Email: leoy.liu@mail.utoronto.ca
 # Date: 26 November 2024
+# Contact: leoy.liu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: -
-# Any other information needed? -
+# Pre-requisites: The `tidyverse` package must be installed
+# Any other information needed? Make sure you are in the `Determinants-of-Wage-Variation-in-Canada` rproj
 
 
 #### Workspace setup ####
 library(tidyverse)
-library(dplyr)
+set.seed(853)
 
 
 #### Simulate data ####
-set.seed(853)
 
 # Set parameters for simulation
 years <- 2000:2019
@@ -22,15 +21,16 @@ education_levels <- c("0 - 8 years", "High school graduate",
                       "Post-secondary certificate or diploma", "Trade certificate or diploma", 
                       "Community college, CEGEP", "University certificate below bachelors degree", 
                       "University degree", "Bachelor's degree", "Above bachelor's degree")
-age_group <- "25-54 years"
+age_groups <- c("15-24 years", "25-54 years", "55 years and over")
+genders <- c("Male", "Female")
 
-# Simulate data
-simulated_data <- expand.grid(Year = years, 
-                              Education_level = education_levels, 
-                              Age_group = age_group)
-
-# Assign education_numeric based on order in education_levels
-simulated_data$Education_numeric <- as.integer(factor(simulated_data$Education_level, levels = education_levels))
+# Create the base structure of the dataset
+simulated_data <- expand.grid(
+  Year = years,
+  Education_level = education_levels,
+  Age_group = age_groups,
+  Gender = genders
+)
 
 # Define the wage rate ranges for each education level
 wage_rate_ranges <- list(
@@ -53,8 +53,9 @@ simulated_data <- simulated_data %>%
 # Reset the order of the rows and remove the row names
 simulated_data <- simulated_data %>%
   ungroup() %>%
-  arrange(Year, Education_level) %>%
-  select(Year, Education_level, Age_group, Avg_hourly_wage_rate, Education_numeric)
+  arrange(Year, Education_level, Age_group, Gender) %>%
[…]

### 2024-12-02 scripts/01-test_simulated_data.R

@@ -1,5 +1,5 @@
 #### Preamble ####
-# Purpose: Tests the structure and validity of the simulated "placeholder" dataset.
+# Purpose: Tests the structure and validity of the simulated dataset.
 # Author: Yuanyi (Leo) Liu
 # Date: 26 November 2024
 # Contact: leoy.liu@mail.utoronto.ca
@@ -7,7 +7,7 @@
 # Pre-requisites: 
 #   - The `tidyverse`, `testthat` packages must be installed and loaded
 #   - 00-simulate_data.R must have been run
-# Any other information needed? Make sure you are in the `Forecasting-the-2024-US-Presidential-Election` rproj
+# Any other information needed? Make sure you are in the `Determinants-of-Wage-Variation-in-Canada` rproj
 
 
 #### Workspace setup ####
@@ -23,28 +23,23 @@ test_that("dataset was successfully loaded", {
 
 
 #### Test data ####
-
-# Check if the dataset has 1000 rows
-test_that("dataset has 1000 rows", {
-  expect_equal(nrow(simulated_data), 1000)
+# Check if the dataset has 1080 rows
+test_that("dataset has 1080 rows", {
+  expect_equal(nrow(simulated_data), 1080)
 })
 
-# Check if the dataset has 11 columns
-test_that("dataset has 11 columns", {
-  expect_equal(ncol(simulated_data), 11)
+# Check if the dataset has 5 columns
+test_that("dataset has 5 columns", {
+  expect_equal(ncol(simulated_data), 5)
 })
 
-# Check if the 'state' column contains only valid US state names
-valid_states <- c("Alabama", "Alaska", "Arizona", "Arkansas", "California", "Colorado", "Connecticut", "Delaware",
-                           "Florida", "Georgia", "Hawaii", "Idaho", "Illinois", "Indiana", "Iowa", "Kansas", "Kentucky",
-                           "Louisiana", "Maine", "Maryland", "Massachusetts", "Michigan", "Minnesota", "Mississippi", 
-                           "Missouri", "Montana", "Nebraska", "Nevada","New Hampshire", "New Jersey", "New Mexico", "New York",
-                           "North Carolina", "North Dakota", "Ohio", "Oklahoma", "Oregon", "Pennsylvania", "Rhode Island", 
-                           "South Carolina", "South Dakota", "Tennessee", "Texas", "Utah", "Vermont", "Virginia", "Washington", 
-                           "West Virginia", "Wisconsin", "Wyoming", "District of Columbia", "National")
-
[…]

### 2024-12-02 scripts/02-clean_data.R

@@ -1,46 +1,0 @@
-#### Preamble ####
-# Purpose: Cleans the raw plane data
-# Author: Yuanyi (Leo) Liu
-# Date: 26 November 2024
-# Contact: leoy.liu@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: raw_data.csv
-# Any other information needed? -
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(arrow)
-
-
-#### Clean data ####
-raw_data <- read_csv("data/raw_data/raw_data.csv")
-
-cleaned_data <-
-  raw_data |>
-  filter(Geography == "Canada",
-         `Type of work` == "Both full- and part-time",
-         Wages == "Average hourly wage rate",
-         `Education level` != "Some high school",
-         `Education level` != "Some post-secondary",
-         `Education level` != "Total, all education levels",
-         `Education level` != "PSE  (5,6,7,8,9))",
-         `Education level` != "No PSE  (0,1,2,3,4)",
-         YEAR >= 2000) |>
-  filter(`Age group` == "25-54 years") |>
-  select(YEAR, `Education level`, `Age group`, `Both Sexes`) |>
-  rename(Year = YEAR, `Avg hourly wage rate` = `Both Sexes`)
-
-education_levels_ordered <- c("0 - 8  years", "High school graduate", 
-                              "Post-secondary certificate or diploma", "Trade certificate or diploma", 
-                              "Community college, CEGEP", "University certificate below bachelors degree", 
-                              "University degree", "Bachelor's degree", "Above bachelor's degree")
-
-cleaned_data <- cleaned_data %>%
-  mutate(`Education level` = factor(`Education level`, levels = education_levels_ordered)) %>%
-  mutate(Education_numeric = as.numeric(`Education level`))
-
-
-#### Save data ####
-write_csv(cleaned_data, "data/analysis_data/analysis_data.csv")
-write_parquet(cleaned_data, "data/analysis_data/analysis_data.parquet")

### 2024-12-02 scripts/02-download_data.R

@@ -1,0 +1,25 @@
+#### Preamble ####
+# Purpose: Downloads and saves the data from FiveThirtyEight.
+# Author: Yuanyi (Leo) Liu, Dezhen Chen, Ziyuan Shen
+# Date: 30 October 2024
+# Contact: leoy.liu@mail.utoronto.ca, dezhen.chen@mail.utoronto.ca, ziyuan.shen@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+# Any other information needed? Make sure you are in the `Forecasting-the-2024-US-Presidential-Election` rproj
+
+
+#### Workspace setup ####
+library(tidyverse)
+
+
+#### Download data ####
+raw_elections_data <-
+  read_csv(
+    file = 
+      "https://data.ontario.ca/dataset/1f14addd-e4fc-4a07-9982-ad98db07ef86/resource/7b325fa1-e9d6-4329-a501-08cdc22a79df/download/v0913_05.csv",
+    show_col_types = TRUE
+  )
+
+
+#### Save data ####
+write_csv(raw_elections_data, "data/01-raw_data/raw_data.csv") 

### 2024-12-02 scripts/03-clean_data.R

@@ -1,0 +1,62 @@
+#### Preamble ####
+# Purpose: Cleans the raw data.
+# Author: Yuanyi (Leo) Liu
+# Date: 26 November 2024
+# Contact: leoy.liu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: 
+#   - The `tidyverse`, `arrow` packages must be installed and loaded
+#   - The raw data must be manually downloaded prior to running this script
+# Any other information needed? Make sure you are in the `Determinants-of-Wage-Variation-in-Canada` rproj
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(arrow)
+
+
+#### Clean data ####
+raw_data <- read_csv("data/01-raw_data/raw_data.csv")
+
+raw_data <- raw_data |>
+  rename(
+    Year = YEAR,
+    Education_level = `Education level`,
+    Age_group = `Age group`,
+    Average_hourly_wages = Wages
+  )
+
+cleaned_data <-
+  raw_data |>
+  filter(Geography == "Canada",
+         `Type of work` == "Both full- and part-time",
+         Average_hourly_wages == "Average hourly wage rate",
+         !Education_level %in% c(
+           "Some high school", "Some post-secondary", 
+           "Total, all education levels", "PSE  (5,6,7,8,9))", 
+           "No PSE  (0,1,2,3,4)"
+         ),
+         Year >= 2000,
+         Age_group %in% c("15-24 years", "25-54 years", "55 years and over")) |>
+  select(Year, Education_level, Age_group, Male, Female) |> 
+  pivot_longer(
+    cols = c(Male, Female), 
+    names_to = "Gender", 
+    values_to = "Average_hourly_wages"
+  )
+
+education_levels_ordered <- c("0 - 8  years", "High school graduate", 
+                              "Post-secondary certificate or diploma", "Trade certificate or diploma", 
+                              "Community college, CEGEP", "University certificate below bachelors degree", 
+                              "University degree", "Bachelor's degree", "Above bachelor's degree")
+
+cleaned_data <- cleaned_data |>
+  mutate(
+    Education_level = factor(Education_level, levels = education_levels_ordered),
+    Education_numeric = as.numeric(Education_level)
+  )
+
+
+#### Save data ####
+write_csv(cleaned_data, "data/02-analysis_data/analysis_data.csv")
+write_parquet(cleaned_data, "data/02-analysis_data/analysis_data.parquet")

### 2024-12-02 scripts/03-test_analysis_data.R

@@ -1,47 +1,0 @@
-#### Preamble ####
-# Purpose: Tests for dataset
-# Author: Yuanyi (Leo) Liu
-# Date: 31 March 2024
-# Contact: leoy.liu@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: 02-data_cleaning.R
-# Any other information needed? -
-
-
-#### Workspace setup ####
-library(tidyverse)
-
-
-#### Test data ####
-# Load dataset
-analysis_data <- read.csv("data/analysis_data/analysis_data.csv", check.names = FALSE)
-
-# Test 1: Check for missing values in simulated_data
-test1_result <- !any(is.na(analysis_data))
-print(paste("Test 1 Passed:", test1_result))
-
-# Test 2: Check for duplicated rows in simulated_data
-test2_result <- !any(duplicated(analysis_data))
-print(paste("Test 2 Passed:", test2_result))
-
-# Test 3: Check that there are exactly 9 unique education levels
-expected_levels <- c("0 - 8  years", "High school graduate", 
-                     "Post-secondary certificate or diploma", "Trade certificate or diploma", 
-                     "Community college, CEGEP", "University certificate below bachelors degree", 
-                     "University degree", "Bachelor's degree", "Above bachelor's degree")
-
-actual_levels <- unique(analysis_data$`Education level`)
-test3_result <- length(actual_levels) == 9 && all(actual_levels %in% expected_levels)
-print(paste("Test 3 Passed:", test3_result))
-
-# Test 4: Check that Education_numeric matches the levels of Education_level
-test4_result <- all(analysis_data$Education_numeric == as.integer(factor(analysis_data$Education_level, levels = education_levels)))
-print(paste("Test 4 Passed:", test4_result))
-
-# Test 5: Check that all years are within the specified range (2000-2019)
-test5_result <- all(analysis_data$Year >= 2000 & analysis_data$Year <= 2019)
-print(paste("Test 5 Passed:", test5_result))
-
-# Test 6: Check that all age groups are "25-54 years"
-test6_result <- all(analysis_data$`Age group` == "25-54 years")
-print(paste("Test 6 Passed:", test6_result))

### 2024-12-02 scripts/04-test_analysis_data.R

@@ -1,0 +1,89 @@
+#### Preamble ####
+# Purpose: Tests structure of the analysis data.
+# Author: Yuanyi (Leo) Liu
+# Date: 26 November 2024
+# Contact: leoy.liu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: 
+#   - The `tidyverse`, `testthat` packages must be installed and loaded
+#   - 03-clean_data.R must have been run
+# Any other information needed? Make sure you are in the `Determinants-of-Wage-Variation-in-Canada` rproj
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(testthat)
+
+analysis_data <- read_csv("data/02-analysis_data/analysis_data.csv")
+
+
+#### Test data ####
+# Test that the dataset has 1080 rows
+test_that("dataset has 1080 rows", {
+  expect_equal(nrow(analysis_data), 1080)
+})
+
+# Test that the dataset has 6 columns
+test_that("dataset has 6 columns", {
+  expect_equal(ncol(analysis_data), 6)
+})
+
+# Test that the 'Education_level' column is character type
+test_that("'Education_level' is character", {
+  expect_type(analysis_data$Education_level, "character")
+})
+
+# Test that the 'Age_group' column is character type
+test_that("'Age_group' is character", {
+  expect_type(analysis_data$Age_group, "character")
+})
+
+# Test that the 'Gender' column is character type
+test_that("'Gender' is character", {
+  expect_type(analysis_data$Gender, "character")
+})
+
+# Test that there are no missing values in the dataset
+test_that("no missing values in dataset", {
+  expect_true(all(!is.na(analysis_data)))
+})
+
+# Test that the 'Education_level' column contains valid education levels
+valid_education_levels <- c("0 - 8  years", "High school graduate", 
+                            "Post-secondary certificate or diploma", "Trade certificate or diploma", 
+                            "Community college, CEGEP", "University certificate below bachelors degree", 
+                            "University degree", "Bachelor's degree", "Above bachelor's degree")
+test_that("'Education_level' column contains valid education levels", {
+  expect_true(all(analysis_data$Education_level %in% valid_education_levels))
+})
+
+# Test that there are no empty strings in 'Education_level', 'Age_group', and 'Gender' columns
+test_that("no empty strings in 'Education_level', 'Age_group', and 'Gender'", {
[…]

### 2024-12-02 scripts/06-model_data.R

@@ -1,28 +1,29 @@
 #### Preamble ####
-# Purpose: Models the dataset with simple linear regression
+# Purpose: Models average hourly wages by using stan_glm.
 # Author: Yuanyi (Leo) Liu
-# Date: 31 March 2024
+# Date: 26 November 2024
 # Contact: leoy.liu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: 02-data_cleaning.R
-# Any other information needed? -
+# Pre-requisites: 
+#   - The `tidyverse`, `rstanarm`, `arrow` packages must be installed and loaded
+#   - 03-clean_data.R must have been run
+# Any other information needed? Make sure you are in the `Determinants-of-Wage-Variation-in-Canada` rproj
 
 
 #### Workspace setup ####
 library(tidyverse)
 library(arrow)
 library(rstanarm)
-library(modelsummary)
 
 
 #### Read data ####
-analysis_data <- read.csv(file = "data/analysis_data/analysis_data.csv")
+analysis_data <- read.csv(file = "data/02-analysis_data/analysis_data.csv")
 
 
 ### Model data ####
-first_model <-
+model <-
   stan_glm(
-    formula = Avg.hourly.wage.rate ~ Education_numeric,
+    formula = Average_hourly_wages ~ Education_level + Age_group + Gender,
     data = analysis_data,
     family = gaussian(),
     prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
@@ -34,6 +35,6 @@ first_model <-
 
 #### Save model ####
 saveRDS(
-  first_model,
+  model,
   file = "models/linear_model.rds"
 )

### 2024-12-02 scripts/07-install_packages.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Install necessary packages
+# Purpose: Install necessary packages.
 # Author: Yuanyi (Leo) Liu
-# Date: 31 March 2024
+# Date: 26 November 2024
 # Contact: leoy.liu@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: -
-# Any other information needed? -
+# Pre-requisites: None
+# Any other information needed? Make sure you are in the `Determinants-of-Wage-Variation-in-Canada` rproj
 
 
 #### Install packages ####

### 2024-12-02 paper/paper.qmd
update model section and reference

@@
 
-The objective of our modeling approach is to quantify the relationship between educational attainment and hourly wages.
-Our analysis employs a Bayesian framework to assess how changes in the level of education correlate with variations in wage rates among Canadian workers aged 25-54.
+Our modeling approach seeks to quantify the relationship between demographic and educational factors and average hourly wages in Canada.
+To achieve this, we employ a linear regression model to examine how predictors such as education level, gender, and age group influence hourly earnings.
+The model is implemented using the `stan_lm` function, with a Gaussian distribution to capture the variability in hourly wages.
 
-In our analysis, we utilized a simple linear regression model, a technique well-suited for examining the relationship between a continuous dependent variable and one or more independent variables.
-Given that our dependent variable, `average hourly wage rate`, is continuous, a normal or Gaussian linear regression model is appropriate.
-This model assumes that the distribution of the wage rate, given the level of education, follows a normal distribution.
-This assumption allows for a straightforward interpretation of the parameters and is commonly used in practice.
-The use of a normal distribution for the error term is a standard choice, as it facilitates the estimation of the model parameters using maximum likelihood estimation.
-The simplicity and interpretability of this model, coupled with its appropriateness for the data at hand, make it a good fit for our study's objectives.
+In this analysis, we focus on predictors that capture key socio-economic and demographic characteristics.
+Specifically, we include `Education_level`, a categorical variable representing individuals' highest level of education; `Gender`, indicating whether the individual identifies as male or female; and `Age_group`, categorized into "15-24 years," "25-54 years," and "55 years and over." These predictors allow us to explore how differences in education, gender, and age intersect to shape wage outcomes.
 
+The model assumes that the average hourly wage, given these predictors, follows a normal distribution.
+This Gaussian assumption simplifies parameter estimation and aligns with standard practices in wage modeling.
+Additionally, we assume moderate priors to avoid overfitting while ensuring interpretability of the coefficients.
+This balanced approach enables us to identify meaningful relationships between the predictors and wages.
 Background details and diagnostics are included in [Appendix -@sec-model-details].
@@
 
+Initially, the `Education_level` variable was grouped into broader categories, such as "Low," "Medium," and "High" education levels, to simplify the analysis.
+This approach aimed to reduce the model’s complexity while capturing general trends in wage variation.
+However, this categorization diminished the model’s ability to detect differences in wages associated with specific education levels.
+For instance, combining "Bachelor’s degree" and "Above bachelor’s degree" into a single category masked the wage attributed to higher education, resulting in a less precise analysis.
+
+An alternative approach was also explored by excluding the `Age_group` variable, based on the hypothesis that its effects might overlap with those of education and gender.
+However, this exclusion led to a poorer model fit, as it failed to account for wage differences across age groups.
+Retaining `Age_group` as a distinct predictor improved the model’s performance, offering a more accurate depiction of how wages vary across life stages.
+
 ## Model set-up
 
-The model predicts hourly wages using the following predictor variables:
+The model predicts the average hourly wage using the following predictor variables:
 
--
+- Education Level (`Education_level`): A categorical variable representing the highest level of formal education attained by an individual. Levels range from "0–8 years" to "Above bachelor's degree."
+
+- Gender (`Gender`): A binary variable indicating whether an individual identifies as "Male" or "Female."
+
+- Age Group (`Age_group`): A categorical variable representing the age ranges of individuals, categorized as "15–24 years", "25–54 years", and "55 years and over".
 
@@
 
-$$
-\begin{aligned}
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \beta_0 +\beta_1x_i\\
-\beta_0 &\sim \mbox{Normal}(0, 2.5) \\
-\beta_1 &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1) \\
-\end{aligned}
-$$
+\begin{align*}
+    y_i \mid \mu_i, \sigma &\sim \text{Normal}(\mu_i, \sigma) \\
+    \mu_i &= \beta_0 + \beta_1 \cdot \text{Education level}_i + \beta_2 \cdot \text{Gender}_i \\
+          &\quad + \beta_3 \cdot \text{Age group}_i + \epsilon_i \\
+    \epsilon_i &\sim \text{Normal}(0, \sigma^2)
+\end{align*}
+
+\textbf{Where:}
+\begin{itemize}
+    \item $\beta_0$ is the intercept term.
+    \item $\beta_1, \beta_2, \beta_3$ are the coefficients for each predictor.
+    \item $\sigma^2$ is the variance of the error term.
+\end{itemize}
 
@@
 
+Existing economic theories and labor market research suggest that education level, age group, and gender notably influence wage outcomes.
+Higher levels of education are associated with greater specialization and skills, which typically result in higher earnings.
+Age reflects work experience and career progression, with older individuals often earning more due to accumulated skills and seniority.
+Gender remains a key determinant, as wage disparities between males and females persist across various labor markets.
+These predictors collectively capture essential socio-economic dimensions that shape wage structures in Canada.
+
+A Bayesian linear regression model was selected because the dependent variable (average hourly wage) is continuous and approximately normally distributed.
+This model is well-suited for assessing the contribution of each predictor to wage outcomes while controlling for the effects of others.
+For instance, the coefficients of education level, age group, and gender directly indicate their respective impacts on average wages.
+
+Further justification for using this model comes from its alignment with the central limit theorem, as the data aggregates wage observations across individuals.
+Additionally, the predictors align with established labor market theories, giving a solid theoretical underpinning to our model.
+The inclusion of priors in the Bayesian framework prevents overfitting, balancing interpretability and predictive accuracy.
+
+A key limitation of this model is that it is entirely trained on the analysis dataset, without splitting into training and testing subsets.
+Each observation in the dataset represents a unique combination of education level, age group, and gender, making it impossible to partition the data without losing critical combinations.
+While this approach ensures that all available data contribute to parameter estimation, it restricts the ability to validate predictions on unseen data.
+To compensate, internal validation methods, such as posterior predictive checks, are employed to evaluate the model’s fit and generalizability.
+Further discussion on this issue is provided in @sec-discussion.
+
 # Results {#sec-result}
+
+## Model Interpretation
+
+## Findings
 
@@
 ## Findings
-
-## Insights on Education and Wages
 
@@
 # Appendix {-}
-
-# Census methodology overview and evaluation
 

### 2024-12-03 other/datasheet/datasheet.qmd
Finalize paper content, update datasheet.

@@
   - Yuanyi (Leo) Liu
-thanks: "Code and data are available at: https://github.com/leoyliu/Analyzing-the-Impact-of-Education-on-Earnings-in-Canada"
+thanks: "Code and data are available at: [Determinants of Wage Variation in Canada](https://github.com/leoyliu/Determinants-of-Wage-Variation-in-Canada)."
 date: today
 date-format: long
-abstract: "This datasheet describes the \"Wages by Education Level\" dataset, which records hourly wage rates across different levels of education in Canada from 2000 to 2019.
-It focuses on individuals aged 25-54 and aims to facilitate analysis of how education impacts wages.
+abstract: "This datasheet provides details about the \"Wages by Education Level\" dataset, which documents hourly wage rates for Canadian workers from 2000 to 2019.
+The dataset includes demographic factors such as age, gender, and education level.
+It is intended to support research on wage disparities and the role of education in shaping economic outcomes.
 The dataset is available under the Open Government Licence - Ontario, allowing for broad use in research and policy discussions.
@@
 
-Extract of the questions from [@gebru2021datasheets]. The data is obtained from the Government of Ontario [@citedata].
+Extract of the questions from "Datasheets for datasets" [@gebru2021datasheets]. The data is obtained from the Government of Ontario [@citedata].
 
@@
 1.  *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
-    -   The dataset was constructed to examine the relationship between education level and average hourly wage rates in Canada between the years 2000 and 2019.
-The intent was to analyze whether educational attainment translates into higher earnings and how this has evolved over time.
+    -   The dataset was developed to investigate the relationship between education level, age, gender, and average hourly wage rates in Canada from 2000 to 2019.
+Its purpose is to assess whether higher educational attainment leads to increased earnings and how this relationship varies across demographic factors and has evolved over time.
 2.  *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
-    -   Yuanyi (Leo) Liu, while studying at the University of Toronto?
+    -   Yuanyi (Leo) Liu, while studying at the University of Toronto.
 3.  *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
@@
 Please provide a description.*
-    -   The dataset represents aggregated wage data by educational attainment for individuals within the age group of 25-54 in Canada. It does not contain multiple types of instances.
+    -   The dataset represents aggregated hourly wage data categorized by education level, age group, and gender for individuals in Canada.
+Each instance corresponds to a unique combination of these factors, capturing variations in wages across demographic and educational categories from 2000 to 2019.
+The dataset does not include multiple types of instances beyond these categories.
 2.  *How many instances are there in total (of each type, if appropriate)?*
-    -   There are 180 instances in the dataset, each representing a unique combination of year and education level.
+    -   There are 1080 instances in the dataset, each representing a unique combination of year, education level, age group, gender, and average hourly wages.
 3.
@@
 4.  *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
-    -   Each instance consists of processed data featuring the year, the level of education, age group, and the average hourly wage rate for the corresponding education level.
+    -  Each instance consists of processed data including the year, education level, age group, gender, and the average hourly wage rate for the corresponding demographic and educational category.
 
 5.  *Is there a label or target associated with each instance? If so, please provide a description.*
-    -   The average hourly wage rate serves as the primary variable of interest, which could be considered the target for analyses examining wage differences across education levels.
+    -  The average hourly wage rate serves as the primary target variable, used to analyze wage differences across education levels, age groups, and genders.
 
@@
 7.  *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
-    -   Relationships are not explicitly stated but can be inferred through the year and education level variables.
+    -   Relationships between instances are not explicitly stated but can be inferred through shared variables such as year, education level, age group, and gender, which allow for comparisons and trend analyses across these dimensions.
 
@@
     -   Preprocessing and cleaning were applied to the dataset, which included filtering specific subsets of data relevant to the study, such as data pertaining to the Canadian geography, average hourly wage rate, and specific age groups.
-The education levels were also ordered numerically to facilitate analysis.
+The education levels were also ordered to facilitate analysis.
 Missing values or incomplete records were not reported; however, any that may have been present were likely excluded during the filtering process.
@@
 1.  *Has the dataset been used for any tasks already? If so, please provide a description.*
-    -   The dataset has been used for an econometric analysis to understand the relationship between education level and average hourly wage rate in Canada within the specified age group and time frame.
+    -   The dataset has been used for an econometric analysis to examine the relationship between education level, age group, gender, and average hourly wage rates in Canada over the specified time frame.
+The analysis aimed to understand how these factors influence earnings and identify trends in wage disparities.
 
 2.  *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
-    -   https://github.com/leoyliu/Analyzing-the-Impact-of-Education-on-Earnings-in-Canada
+    -   https://github.com/leoyliu/Determinants-of-Wage-Variation-in-Canada
 
@@
 Is there anything a dataset consumer could do to mitigate these risks or harms?*
-    -   The way the dataset has been collected and processed should be considered when conducting future research to ensure that the socioeconomic context is taken into account, especially considering the dataset only covers a specific age group and may not generalize across all demographics.
-Future users should ensure that their research questions and methods are compatible with the dataset's limitations to avoid skewed interpretations.
+    -   The dataset's composition and preprocessing choices should be considered in future research.
+It focuses on specific age groups, excludes some education categories, and does not cover all demographics, which may limit its generalizability.
+Additionally, the aggregation of data might obscure variations within subpopulations.
+Users should align their research questions with the dataset's scope and limitations and contextualize findings to avoid biased or oversimplified interpretations.
+Incorporating supplementary data sources or validating results against broader datasets can mitigate these risks.
 

### 2024-12-03 paper/paper.qmd
Finalize paper content, update datasheet.

@@
 ```
+
+\newpage
 
@@
     booktabs = TRUE
-  )
+  ) %>%
+  kable_styling(font_size = 10, full_width = FALSE)
 ```
@@
 
-@fig-avg-wage illustrates the trend of average hourly wages in Canada from 2000 to 2019.
+Using package `ggplot2` [@citeggplot], @fig-avg-wage presents the trend of average hourly wages in Canada from 2000 to 2019.
 Each dot represents the average wage for a specific year, and the overall shows a upward trend.
@@
 A key limitation of this model is that it is entirely trained on the analysis dataset, without splitting into training and testing subsets.
-Each observation in the dataset represents a unique combination of education level, age group, and gender, making it impossible to partition the data without losing critical combinations.
+Each observation in the dataset represents a unique combination of education level, age group, and gender, making it impossible to partition the data without losing combinations.
 While this approach ensures that all available data contribute to parameter estimation, it restricts the ability to validate predictions on unseen data.
@@
 
-## Model Interpretation
+@sec-result examines the relationship between education level, gender, and age group with respect to average hourly wages in Canada.
+Using the analysis dataset, we apply a Bayesian linear regression model to identify the key factors that influence hourly wages.
+Below, we present the results of our model, visualizations, and key takeaways.
 
-## Findings
+## Model results and interpretation
+
+The linear regression model built on the analysis dataset estimated the effects of education level, age group, and gender on average hourly wages.
+The results are summarized in @tbl-summary, with a focus on the coefficients and their implications.
+The intercept, estimated at 5.758, represents the baseline hourly wage for females aged 15-24 years with an education level of "0–8 years."
+
+```{r}
+#| label: tbl-summary
+#| tbl-cap: "Summary of regression model coefficients examining the relationship between education level, age group, gender, and average hourly wages in Canada."
+#| echo: false
+#| warning: false
+
+# Load the model
+model <- readRDS(file = here::here("models/linear_model.rds"))
+
+modelsummary(
+  list(
+    "coefficient" = model
+  )
+)
+```
+
+### Education Level
+
+Education level has a strong positive impact on hourly wages.
+Individuals with "Above bachelor’s degree" earn approximately 13.844 CAD more per hour compared to those with "0–8 years" of education.
+Similarly, holding a bachelor’s degree is associated with an increase of 10.726 CAD per hour, and having a university degree contributes an additional 11.740 CAD.
+
+### Age Group
+
+Hourly wages vary considerably across age groups.
+Individuals aged 25–54 years earn 9.630 CAD more per hour than those in the baseline group (15–24 years).
+Those in the 55+ age group experience an even higher increase of 10.546 CAD, reflecting the wage advantages linked to experience and seniority.
+
+### Gender
+
+The analysis highlights gender disparities in wages, with males earning 3.618 CAD more per hour than females on average.
+This finding reflects ongoing gender wage gaps, even after accounting for education and age group.
+
+Overall, the model explains a substantial portion of the variance in hourly wages, with an $R^2$ of 0.781, indicating that 78.1% of the variation is captured by the predictors.
+The adjusted $R^2$ of 0.778 supports the accuracy of the model, while the RMSE value of 3.45 suggests a reasonable level of error in predicting wages.
+These results align with economic theories that emphasize education as a key driver of earning potential, age as a proxy for experience, and persistent gender wage gaps in labor markets.
+A complete overview of the model and diagnostic details is available in [Appendix -@sec-model-details].
+
+## Visualization of Results
+
+The relationship between education level, gender, and age group with average hourly wages is illustrated in @fig-visualize-model.
+Each panel represents a distinct age group: "15–24 years," "25–54 years," and "55 years and over." Within each panel, points are color-coded by gender.
+
+```{r fig.width=10, fig.height=7}
+#| label: fig-visualize-model
+#| fig-cap: "Average hourly wages by education level, gender, and age group in Canada.
+Panels represent age groups, with points color-coded by gender.
+Dashed lines show linear trends, highlighting wage differences across education levels, genders, and age groups."
+#| echo: false
+#| warning: false
+
+cleaned_data |> 
+   ggplot(aes(x = reorder(Education_level, Average_hourly_wages, FUN = median), 
+             y = Average_hourly_wages, 
+             colour = Gender)) + 
+  geom_point(alpha = 0.5) + 
+  geom_smooth(
+    aes(group = Gender),
+    method = "lm",
+    se = TRUE,
+    color = "black",
+    linetype = "dashed",
+    formula = y ~ x
+  ) + 
+  scale_colour_manual(
+    values = c("Female" = "#FF6F61", "Male" = "#6495ED")
+  ) +
+  labs(
+    x = "Education Level",
+    y = "Average Hourly Wage",
+    colour = "Gender"
+  ) + 
+  facet_wrap(~Age_group, ncol = 3) +
+  theme_classic() + 
+  theme(
+    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 9),
+    axis.text.y = element_text(size = 9),
+    axis.title.x = element_text(size = 12),
+    axis.title.y = element_text(size = 12),
+    legend.position = "right",
+    legend.title = element_text(hjust = 0.5)
+  ) + 
+  guides(colour = guide_legend(title = "Gender", ncol = 1))
+```
+
+@fig-visualize-model presents that wages consistently increase with higher levels of education across all age groups.
+This trend is evident in the steepness of the linear trendlines, showing that the wage benefits associated with university degrees and qualifications above a bachelor’s degree.
+The gender wage gap is visible in all panels, with males earning higher wages than females at nearly every education level.
+Notably, the gap widens at higher education levels, suggesting that gender disparities are more significant as qualifications increase.
+
+It also highlights differences in wages across age groups.
+Older individuals, particularly those aged 55 and over, earn higher wages compared to younger cohorts, reflecting the influence of experience and career progression.
+Additionally, the intersection of gender, age, and education reveals further disparities.
+For example, males with higher education levels in older age groups consistently exhibit the highest wages, while younger females with lower education levels earn the least.
 
@@
 
-## Findings
+## Key Findings
+
+### The Impact of Education on Wages
+
+Higher levels of education are strongly associated with increased hourly wages.
+Individuals with a bachelor’s degree or higher earn more than those with lower educational attainment.
+This trend reflects the demands of a knowledge-based economy, where employers value specialized qualifications and technical expertise.
+It underscores the role of education in enhancing skills and productivity, which directly translates to higher earnings.
+
+Moreover, the wage premium for higher education shows the importance of equitable access to education.
+Rising costs of tuition and other barriers could exacerbate socioeconomic inequalities, as those unable to access higher education may remain confined to lower-paying jobs.
+Policymakers should consider expanding access to affordable education to ensure broader participation in higher-paying sectors.
+
+### Age and Experience in Wage Growth
+
+The results reveal a clear wage advantage for middle-aged and older workers compared to younger individuals.
+Workers aged 25–54 years and 55+ years earn substantially more than their counterparts aged 15–24 years.
+This trend is likely attributable to experience, seniority, and the accumulation of skills over time, which enhance an individual’s productivity and value in the labor market.
+
+In practical terms, this trend reflects a lifecycle wage profile, where earnings typically peak during mid-career and stabilize or slightly decline as workers approach retirement.
+This understanding is important for retirement planning and addressing income inequality across life stages.
+Employers could also consider strategies to support younger workers, such as offering training programs and mentorship opportunities, to accelerate their skill development and income growth.
+
+### Persistent Gender Wage Disparities
+
+The analysis reveals a consistent wage disparity between males and females, with males earning much more across all education levels and age groups.
+This persistent wage gap suggests systemic inequities in how labor markets reward work.
+Factors contributing to this disparity may include occupational segregation, differences in work hours, career interruptions, and implicit biases in hiring and promotion processes.
+
+Addressing this gap requires targeted interventions.
+Initiatives such as wage transparency, mentorship programs for women, and policies promoting work-life balance could help bridge the divide.
+Beyond fairness, closing the gender wage gap has broader economic implications, as it could boost household incomes and enhance overall productivity in the economy.
 
 ## Limitations and Future Research Directions
+
+This study relies on a dataset that represents unique combinations of education level, age group, and gender, limiting the use of traditional validation methods like splitting data into training and testing sets.
+Future research could explore alternative approaches, such as bootstrapping or cross-validation, to enhance model.
+
+Additionally, the dataset focuses on a specific subset of the population and does not account for variables such as geographic location, race, or immigration status.
+These factors likely intersect with education, age, and gender to influence wages, and their inclusion could provide a more detailed understanding of wage disparities.
+
+The findings emphasize the economic returns of education, highlighting the need for investment in accessible, high-quality educational systems.
+The relationship between age and wages underscores the importance of workforce development and retirement planning.
+Additionally, addressing gender wage disparities requires systemic changes to ensure equitable rewards for labor.
+
+Future studies could investigate the causal mechanisms behind these trends using longitudinal data or natural experiments.
+Evaluating the effectiveness of policies, such as wage transparency laws or educational subsidies, could provide actionable understanding.
+Further research into the intersection of race, immigration status, and other factors with wages would deepen our analysis of labor market dynamics.
 
@@
 
+## Overview
+
+This survey is designed to investigate how demographic and educational factors influence average hourly wages in Canada.
+With a hypothetical budget of $50,000, the survey methodology aims to capture detailed and representative data from working-age individuals across diverse demographics.
+The survey focuses on three key variables: education level, age group, and gender, and it seeks to ensure accuracy, minimize bias, and provide coverage of the Canadian workforce.
+
+## Sampling approach
+
+The survey employs a stratified random sampling strategy to ensure that the collected data represents the Canadian working population.
+This approach helps capture wage patterns across various demographic groups.
+
+### Stratification variables
+
+* Age Group: 15–24 years, 25–54 years, 55 years and over
+* Gender: Male, Female, and Non-binary
+* Education Level: 0 – 8 years, High school graduate, Trade certificate or diploma, Post-secondary certificate or diploma, Community college, University certificate below bachelor's degree, University degree, Bachelor's degree, Above bachelor's degree
+* Employment Type: Full-time workers, Part-time workers
+* Race: White, Black, Indigenous, Asian, Hispanic/Latino, Other
+* Hourly Income:  <\$15,  \$15-\$29, \$30-\$44, \$45-\$59, >\$60
+
+### Sample Size
+
+The survey targets 20,000 respondents across Canada.
+This sample size ensures a balance between budget constraints and statistical reliability, with an expected margin of error of ±2.5% at a 95% confidence level.
+
+## Recruitment Strategy
+
+To ensure a diverse and representative sample, the survey incorporates two recruitment strategies:
+
+**Online survey panels:**
+
+* Partner with reputable online survey platforms, such as SurveyMonkey or YouGov, which have access to extensive databases of respondents.
+* Validate participant demographics to match the stratification variables.
+
+**Social media advertising:**
+
+* Use platforms such as Facebook, LinkedIn, and Instagram to reach underrepresented groups, including younger individuals, minority groups, and non-binary respondents.
+* Tailor advertisements to target specific demographics and encourage participation.
+
+**Incentives:** Participants will be entered into a raffle for gift card prizes, ensuring high participation rates.
+
+## Data validation and quality control
+
+To maintain the integrity of the data, we will take the following measures:
+
+* **Attention checks:** At least two 'attention check' questions will be included in the survey to identify inattention or robots.
+* **Duplicate detection:** Use email or IP-based verification to prevent duplicate responses. Ensure that each respondent provides only one response set to keep the dataset unique.
+* **Response time monitoring:** Track the time it takes respondents to complete surveys. Responses with apparently faster-than-average completion times are labelled as potentially low-quality.
+* **Data consistency check:** Analyses the consistency of responses, especially between related questions. Inconsistent answers will be labelled for review.
+* **Panel partner validation:** For respondents recruited through an online survey panel, rely on the panel provider's in-built validation systems.
+These systems use various quality checks, such as identity verification and response consistency, which ensure high-quality data from panel participants.
+
+## Budget allocation
+
+* Online Panel Recruitment: $25,000
+* Social Media Advertising: $15,000
+* Participant Incentives: $5,000
+* Data Cleaning and Validation: $5,000
+
+## Survey implementation
+The survey will be available via Google Forms to ensure broad accessibility and easy data collection.
+The link to the survey will be distributed via email invitations to panellists and targeted social media adverts.
+To ensure maximum reach and diversity, multiple channels will be used, including Facebook, Instagram, and other social media platforms.
+The link to the survey is [here](https://forms.gle/UkkD4Xt6FaJeusu47).
+
+**Distribution plan:**
+
+* **Panel invitations:** Members of established survey panels (e.g. Lucid, YouGov) will receive email invitations to participate in the survey.
+* **Social media activities:** Targeted adverts will be placed on social media platforms to reach under-represented groups, especially young voters and national minority groups.
+* **Follow-up reminders:** An automated reminder email will be sent to participants who have not completed the survey to maximize response rates.
+* **Survey monitoring:** Response rates will be closely monitored throughout the data collection period and sampling methods will be adjusted if certain population quotas are not reached.
+* **Accessibility:** The survey will be mobile-friendly, which will allow respondents to complete the survey on any device.
+
+## Survey structure
+
+**Survey introduction:**
+
+Welcome!
+
+We are conducting a survey to understand how demographic and educational factors influence hourly wages in Canada.
+Your participation is important, and your responses will contribute to valuable research on wage distribution across various groups.
+
+Please note:
+
+* All responses are confidential and will be used for research purposes only.
+* This survey will take approximately 10 minutes to complete.
+* Your participation is entirely voluntary, and you may withdraw at any time.
+* All questions marked with an asterisk are required.
+* There are no right or wrong answers, we appreciate your honest opinions.
+
+If you have any questions or concerns, please feel free to contact our research team at [liuyuanyi1208@gmail.com](mailto:liuyuanyi1208@gmail.com).
+
+As a thank-you for your participation, you will be entered into a raffle to win a gift card. We deeply appreciate the time and effort you contribute to this study.
+
+**Survey question:**
+
+**Part 1: Screening for eligibility**
+
+1.Are you currently employed?\
+* Yes / No\
+2.Are you aged 15 years or older?\
+* Yes / No\
+3.Do you currently reside in Canada?\
+* Yes / No
+
+**Part 2: Demographics**
+
+4.What is your age group?\
+* 15–24 years / 25–54 years / 55+\
+5.What is your gender?\
+* Male / Female / Non-binary / Prefer not to say\
+6.Which of the following best describes your race or ethnicity?
+* White / Black / Indigenous / Asian / Hispanic/Latino / Other\
+7.What is your highest level of education?\
+* 0–8 years / High school graduate / Trade certificate or diploma / Post-secondary certificate or diploma / Community college, CEGEP / University certificate below bachelor's degree / Bachelor's degree / University degree / Above bachelor's degree
+
+**Part 3: Employment details**
+
+8.Are you employed full-time or part-time?\
+* Full-time / Part-time\
+9.In which industry or job sector do you work?\
+* [Open-ended response]\
+10.How many hours do you typically work per week?\
+* [Numeric response]
+
+**Part 4: Wage**
+
+11.What is your average hourly income before taxes?\
+* [Numeric response]\
+12.Does your wage include additional benefits or bonuses?\
+* Yes / No\
+13.Do you believe your education level has impacted your wages?\
+* Yes / No / Not Sure
+
+**Part 5: Verify and Consent**
+
+14.Please select 'Agree' to verify that you are paying attention.\
+* Agree / Disagree\
+15.Do you agree to participate in this survey? Your responses will be kept confidential and used only for research purposes.\
+* Yes / No
+
+**End Part:**
+
+Thank you for your participation!
+
+Your responses are valuable to us in understanding wage trends in Canada.
+If you have any questions or would like more information about this research, please feel free to contact us at [liuyuanyi1208@gmail.com](mailto:liuyuanyi1208@gmail.com).
+
 # Data Manipulation and Cleaning {#sec-data-cleaning}
 
-During the data cleaning phase, the R packages `tidyverse` [@citetidyverse], `dplyr` [@citedplyr], and `arrow` [@citearrow] were utilized.
+During the data cleaning phase, the R packages `tidyverse` [@citetidyverse], `dplyr` [@citedplyr], and `arrow` [@citearrow] were used.
 The raw data was imported using `read_csv` from the `tidyverse` [@citetidyverse] package.
@@
 
-The analysis was specifically targeted at the demographic group aged between 25 and 54 years, as this range represents a prime working-age population.
-Columns important to the analysis were chosen and renamed for better readability and straightforward reference in subsequent analytical procedures.
-This process included renaming the columns for the year and average hourly wage rate to 'Year' and 'Avg hourly wage rate' respectively.
+The analysis focused on data from 2000 onward, ensuring the timeframe was appropriate for the study.
+Age groups were restricted to "15–24 years," "25–54 years," and "55 years and over," representing distinct workforce demographics.
+Columns were renamed for clarity, including Year, Education_level, and Average_hourly_wages, enabling easier reference in subsequent analyses.
 
-Education levels were organized in a logical order ranging from "0 - 8 years" of education to "Above bachelor's degree" using the `factor` function to convert them into a categorical variable with a specified level order.
-This categorization was further enhanced by creating an accompanying numeric variable that mapped these ordered education levels to integers, thus facilitating quantitative analysis.
+To facilitate gender comparisons, the dataset was reshaped using the `pivot_longer` function, combining the Male and Female wage columns into a single column labeled Gender, with corresponding wage data stored under Average_hourly_wages.
+This restructuring allowed for straightforward comparisons across gender groups while maintaining a tidy dataset.
+
+Education levels were ordered logically, from "0–8 years" to "Above bachelor’s degree," using the `reorder` function.
+This ensured consistent interpretation in visualizations and modeling, highlighting wage trends across increasing educational attainment levels.
 
 The final step involved saving the cleaned and structured data.
-This was done using the `write_csv` function to generate a CSV file for broad compatibility and the `write_parquet` function from the `arrow` [@citearrow] package for a more compressed and efficient file format, both of which were stored in the `data/analysis_data` directory.
-
+This was done using the `write_csv` function to generate a CSV file for broad compatibility and the `write_parquet` function from the `arrow` [@citearrow] package for a more compressed and efficient file format, both of which were stored in the `data/02-analysis_data` directory.
 In all figures and tables, the library `here` [@citehere] was used to ensure that the file path should be accessible in all directories.
@@
 
+In @fig-ppcheck, we implement a posterior predictive check, which displays the overlap between the observed data (denoted by $y$) and the replicated data generated from the model (denoted by $y_{rep}$).
+In a well-fitting model, the distribution of these replicated data should resemble the distribution of the observed data.
+Here, the replicated lines mainly follow the shape and central tendency of the observed data’s density.
+This similarity indicates that the model is capturing the overall pattern of the data reasonably well.
+Deviations between the replicated and observed lines would indicate areas where the model might not be accurately capturing the data structure.
+
+```{r}
+#| eval: true
+#| echo: false
+#| message: false
+#| warning: false
+#| label: fig-ppcheck
+#| fig-cap: "Comparison of Observed Data Density (dark line) and Model-Generated Replications (light lines) shows close alignment, suggesting the model captures the main distributional characteristics of the data."
+
+pp_check(model) +
+  theme_classic() +
+  theme(legend.position = "right")
+```
+
 ## Diagnostics
+
+```{r}
+#| echo: false
+#| eval: true
+#| message: false
+#| warning: false
+#| label: fig-rhat
+#| fig-cap: "All parameters have R-hat values at or below 1.05, indicating strong convergence of the MCMC chains and reliable parameter estimates."
+
+plot(model, "rhat")
+```
+
+@fig-rhat is a visualization of the Gelman-Rubin diagnostic, often denoted as R-hat, which assesses the convergence of a Bayesian model’s Markov Chain Monte Carlo (MCMC) samples.
+It shows the values of for various parameters in the model.
+An value close to 1 indicates good convergence, meaning the chains for each parameter have mixed well and are sampling from the same posterior distribution.
+In @fig-rhat, all parameters have an value at or below 1.05, suggesting that the model has achieved adequate convergence and the estimates are reliable.
 

### 2024-12-03 scripts/02-clean_data.R

@@ -50,12 +50,6 @@ education_levels_ordered <- c("0 - 8  years", "High school graduate",
                               "Community college, CEGEP", "University certificate below bachelors degree", 
                               "University degree", "Bachelor's degree", "Above bachelor's degree")
 
-cleaned_data <- cleaned_data |>
-  mutate(
-    Education_level = factor(Education_level, levels = education_levels_ordered),
-    Education_numeric = as.numeric(Education_level)
-  )
-
 
 #### Save data ####
 write_csv(cleaned_data, "data/02-analysis_data/analysis_data.csv")

### 2024-12-03 scripts/02-download_data.R

@@ -1,25 +1,0 @@
-#### Preamble ####
-# Purpose: Downloads and saves the data from FiveThirtyEight.
-# Author: Yuanyi (Leo) Liu, Dezhen Chen, Ziyuan Shen
-# Date: 30 October 2024
-# Contact: leoy.liu@mail.utoronto.ca, dezhen.chen@mail.utoronto.ca, ziyuan.shen@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: None
-# Any other information needed? Make sure you are in the `Forecasting-the-2024-US-Presidential-Election` rproj
-
-
-#### Workspace setup ####
-library(tidyverse)
-
-
-#### Download data ####
-raw_elections_data <-
-  read_csv(
-    file = 
-      "https://data.ontario.ca/dataset/1f14addd-e4fc-4a07-9982-ad98db07ef86/resource/7b325fa1-e9d6-4329-a501-08cdc22a79df/download/v0913_05.csv",
-    show_col_types = TRUE
-  )
-
-
-#### Save data ####
-write_csv(raw_elections_data, "data/01-raw_data/raw_data.csv") 

### 2024-12-03 scripts/03-test_analysis_data.R

@@ -6,7 +6,7 @@
 # License: MIT
 # Pre-requisites: 
 #   - The `tidyverse`, `testthat` packages must be installed and loaded
-#   - 03-clean_data.R must have been run
+#   - 02-clean_data.R must have been run
 # Any other information needed? Make sure you are in the `Determinants-of-Wage-Variation-in-Canada` rproj
 
 
@@ -23,9 +23,9 @@ test_that("dataset has 1080 rows", {
   expect_equal(nrow(analysis_data), 1080)
 })
 
-# Test that the dataset has 6 columns
-test_that("dataset has 6 columns", {
-  expect_equal(ncol(analysis_data), 6)
+# Test that the dataset has 5 columns
+test_that("dataset has 5 columns", {
+  expect_equal(ncol(analysis_data), 5)
 })
 
 # Test that the 'Education_level' column is character type

### 2024-12-03 scripts/04-exploratory_data_analysis.R

@@ -1,0 +1,106 @@
+#### Preamble ####
+# Purpose: Exploratory data analyis on cleaned data.
+# Author: Yuanyi (Leo) Liu
+# Date: 26 November 2024
+# Contact: leoy.liu@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: 
+#   - The `tidyverse`, `rstanarm` packages must be installed and loaded
+#   - 02-clean_data.R must have been run
+# Any other information needed? Make sure you are in the `Forecasting-the-2024-US-Presidential-Election` rproj
+
+
+#### Workspace setup ####
+library(tidyverse)
+library(rstanarm)
+
+#### Read data ####
+analysis_data <- read_csv("data/02-analysis_data/analysis_data.csv")
+
+# General overview of the dataset
+glimpse(analysis_data)
+summary(analysis_data)
+
+
+#### Summary Statistics ####
+# Summary of numeric columns
+analysis_data %>%
+  select_if(is.numeric) %>%
+  summary()
+
+# Count unique values in categorical columns
+analysis_data %>%
+  select_if(is.character) %>%
+  summarise_all(~ n_distinct(.))
+
+#### Missing Values Analysis ####
+# Check for missing values in each column
+analysis_data %>%
+  summarise_all(~ sum(is.na(.)))
+
+
+#### Univariate Analysis ####
+# Distribution of Average Hourly Wages
+ggplot(analysis_data, aes(x = Average_hourly_wages)) +
+  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
+  labs(title = "Distribution of Average Hourly Wages", x = "Average Hourly Wages (CAD)", y = "Count")
+
+# Distribution of Education Levels
+ggplot(analysis_data, aes(x = reorder(Education_level, Average_hourly_wages, median))) +
+  geom_bar(fill = "lightgreen") +
+  coord_flip() +
+  labs(title = "Distribution of Education Levels", x = "Education Level", y = "Count")
+
+# Distribution of Age Groups
+ggplot(analysis_data, aes(x = Age_group)) +
+  geom_bar(fill = "coral") +
+  labs(title = "Distribution of Age Groups", x = "Age Group", y = "Count")
+
+# Distribution of Gender
+ggplot(analysis_data, aes(x = Gender)) +
+  geom_bar(fill = "purple") +
[…]

### 2024-12-03 scripts/05-exploratory_data_analysis.R

@@ -1,113 +1,0 @@
-#### Preamble ####
-# Purpose: Exploratory data analyis on cleaned data.
-# Author: Yuanyi (Leo) Liu
-# Date: 26 November 2024
-# Contact: leoy.liu@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: 
-#   - The `tidyverse`, `rstanarm` packages must be installed and loaded
-#   - 03-clean_data.R must have been run
-# Any other information needed? Make sure you are in the `Forecasting-the-2024-US-Presidential-Election` rproj
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(rstanarm)
-
-#### Read data ####
-analysis_data <- read_csv("data/02-analysis_data/analysis_data.csv")
-
-# General overview of the dataset
-glimpse(analysis_data)
-summary(analysis_data)
-
-
-#### Summary Statistics ####
-# Summary of numeric columns
-analysis_data %>%
-  select_if(is.numeric) %>%
-  summary()
-
-# Count unique values in categorical columns
-analysis_data %>%
-  select_if(is.character) %>%
-  summarise_all(~ n_distinct(.))
-
-#### Missing Values Analysis ####
-# Check for missing values in each column
-analysis_data %>%
-  summarise_all(~ sum(is.na(.)))
-
-
-#### Univariate Analysis ####
-# Distribution of numeric_grade
-ggplot(analysis_data, aes(x = numeric_grade)) +
-  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
-  labs(title = "Distribution of Poll Numeric Grades", x = "Numeric Grade", y = "Count")
-
-# Distribution of transparency_score
-ggplot(analysis_data, aes(x = transparency_score)) +
-  geom_histogram(bins = 20, fill = "lightgreen", color = "black") +
-  labs(title = "Distribution of Poll Transparency Scores", x = "Transparency Score", y = "Count")
-
-# Distribution of sample_size
-ggplot(analysis_data, aes(x = sample_size)) +
-  geom_histogram(bins = 30, fill = "coral", color = "black") +
-  labs(title = "Distribution of Poll Sample Sizes", x = "Sample Size", y = "Count") +
-  scale_x_log10()  # Log scale for better visualization
-
-# Distribution of percent (poll result percentage)
-ggplot(analysis_data, aes(x = pct)) +
-  geom_histogram(bins = 30, fill = "purple", color = "black") +
[…]

### 2024-12-03 scripts/05-model_data.R

@@ -6,7 +6,7 @@
 # License: MIT
 # Pre-requisites: 
 #   - The `tidyverse`, `rstanarm`, `arrow` packages must be installed and loaded
-#   - 03-clean_data.R must have been run
+#   - 02-clean_data.R must have been run
 # Any other information needed? Make sure you are in the `Determinants-of-Wage-Variation-in-Canada` rproj
 
 

### 2024-12-03 scripts/06-install_packages.R

@@ -17,3 +17,6 @@ install.packages("here")
 install.packages("ggplot2")
 install.packages("kableExtra")
 install.packages("modelsummary")
+install.packages("janitor")
+install.packages("knitr")
+install.packages("testthat")

### 2024-12-04 other/datasheet/datasheet.qmd
Finalize paper, update sketches and datasheet

@@
 Its purpose is to assess whether higher educational attainment leads to increased earnings and how this relationship varies across demographic factors and has evolved over time.
+
 2.  *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
     -   Yuanyi (Leo) Liu, while studying at the University of Toronto.
+
 3.  *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
     -   No applicable.
+
 4.  *Any other comments?*
@@
 The dataset does not include multiple types of instances beyond these categories.
+
 2.  *How many instances are there in total (of each type, if appropriate)?*
     -   There are 1080 instances in the dataset, each representing a unique combination of year, education level, age group, gender, and average hourly wages.
+
 3.
@@
 8.  *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
-    -   There are no recommended data splits mentioned, as the data appears to be used for comprehensive temporal analysis rather than predictive modeling that would require such splits.
+    -   There are no recommended data splits mentioned, as the data appears to be used for temporal analysis rather than predictive modeling that would require such splits.
 
@@
 3.  *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
-    -   As the dataset was derived from a comprehensive administrative record, the sampling strategy is not applicable. The data represents a census of the information available rather than a sample.
+    -   As the dataset was derived from an administrative record, the sampling strategy is not applicable. The data represents a census of the information available rather than a sample.
 

### 2024-12-04 paper/paper.qmd
Finalize paper, update sketches and datasheet

@@
 @fig-education-level presents the distribution of average hourly wages in Canada from 2000 to 2019 across different education levels.
-Higher education levels, such as "University degree" and "Above bachelor's degree," are associated with higher median wages, while lower education levels, like "0–8 years," show significantly lower wages.
+Higher education levels, such as "University degree" and "Above bachelor's degree," are associated with higher median wages, while lower education levels, like "0–8 years," show lower wages.
 The variability within each education level, shown by the spread of the boxes and whiskers, indicates differences in earning potential within these groups.
@@
 The jittered points represent individual wage observations, further emphasizing the overlap and differences in wage ranges.
-While both distributions have a similar shape, the median wage for males is noticeably higher than that for females, reflecting a persistent gender wage gap over the period analyzed.
+While both distributions have a similar shape, the median wage for males is higher than that for females, reflecting a persistent gender wage gap over the period analyzed.
 
@@
 Older individuals, particularly those aged 55 and over, earn higher wages compared to younger cohorts, reflecting the influence of experience and career progression.
-Additionally, the intersection of gender, age, and education reveals further disparities.
+Additionally, the intersection of gender, age, and education shows further disparities.
 For example, males with higher education levels in older age groups consistently exhibit the highest wages, while younger females with lower education levels earn the least.
@@
 
-The analysis reveals a consistent wage disparity between males and females, with males earning much more across all education levels and age groups.
+The analysis presents a consistent wage disparity between males and females, with males earning much more across all education levels and age groups.
 This persistent wage gap suggests systemic inequities in how labor markets reward work.

### 2024-12-04 scripts/04-exploratory_data_analysis.R

@@ -7,7 +7,7 @@
 # Pre-requisites: 
 #   - The `tidyverse`, `rstanarm` packages must be installed and loaded
 #   - 02-clean_data.R must have been run
-# Any other information needed? Make sure you are in the `Forecasting-the-2024-US-Presidential-Election` rproj
+# Any other information needed? Make sure you are in the `Determinants-of-Wage-Variation-in-Canada` rproj
 
 
 #### Workspace setup ####

### 2024-12-14 paper/paper.qmd
fix prose

@@
 #| label: fig-education-level
-#| fig-cap: "Average hourly wages in Canada by education level from 2000 to 2019.
-Each box represents the wage distribution for a specific education level, with higher education levels corresponding to higher median wages."
+#| fig-cap: "Average hourly wages in Canada by education level from 2000 to 2019. Higher education levels corresponding to higher median wages."
 #| echo: false
@@
 #| label: fig-gender
-#| fig-cap: "Average hourly wages in Canada by gender from 2000 to 2019.
-Each violin plot represents the wage distribution for males and females, with males showing higher median wages and a broader range of earnings compared to females."
+#| fig-cap: "Average hourly wages in Canada by gender from 2000 to 2019. Males showing higher median wages and a broader range of earnings compared to females."
 #| echo: false
@@
 #| label: fig-age
-#| fig-cap: "Average hourly wages in Canada by age group from 2000 to 2019.
-Each curve represents the wage distribution for a specific age group, with older age groups showing higher average wages and broader distributions compared to younger groups."
+#| fig-cap: "Average hourly wages in Canada by age group from 2000 to 2019. Older age groups showing higher average wages and broader distributions compared to younger groups."
 #| echo: false
@@
 #| label: fig-visualize-model
-#| fig-cap: "Average hourly wages by education level, gender, and age group in Canada.
-Panels represent age groups, with points color-coded by gender.
-Dashed lines show linear trends, highlighting wage differences across education levels, genders, and age groups."
+#| fig-cap: "Average hourly wages by education level, gender, and age group in Canada. Wages increase with higher levels of education across all age groups."
 #| echo: false
@@
 We are conducting a survey to understand how demographic and educational factors influence hourly wages in Canada.
-Your participation is important, and your responses will contribute to valuable research on wage distribution across various groups.
+Your participation is important, and your responses will contribute to the research on wage distribution across various groups.
 
@@
 
-Your responses are valuable to us in understanding wage trends in Canada.
+Your responses are important to us in understanding wage trends in Canada.
 If you have any questions or would like more information about this research, please feel free to contact us at [liuyuanyi1208@gmail.com](mailto:liuyuanyi1208@gmail.com).
